---
title: "Weibull AFT and Cure Models with vldaft"
subtitle: "Location, scale, shape, and cure-rate effects on survival"
author: "Keaven Anderson"
date: today
format:
  revealjs:
    theme: simple
    slide-number: c/t
    show-slide-number: all
    chalkboard: false
    incremental: false
    smaller: true
    width: 1280
    height: 900
    fig-width: 8
    fig-height: 4.5
    fig-align: center
    code-line-numbers: false
    auto-stretch: false
    center: false
    embed-resources: true
    css: custom.css
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include = FALSE}
library(gglite)
```

## Outline

1. The Weibull AFT model and its parameterization
2. How **location** ($\mu$) shifts the survival curve
3. How **scale** ($\sigma$) reshapes the survival curve
4. The VLDAFT extension: letting $\sigma$ depend on $\mu$
5. Anderson (1991): Framingham Heart Study application
6. Cure models: adding a survival plateau
7. Cure rate + shape interaction

---

## The Weibull AFT parameterization

For event time $T$:

$$
\log T = \mu + \sigma\,\varepsilon, \qquad \varepsilon \sim \text{Gumbel}
$$

The survival function is:

$$
S(t) = \exp\!\left\{-\exp\!\left(\frac{\log t - \mu}{\sigma}\right)\right\}
$$

- $\mu$ = **location** on the log-time scale (shifts the curve left/right)
- $\sigma$ = **scale** (controls the spread / shape of the hazard)
- Shape parameter $\alpha = 1/\sigma$: when $\alpha > 1$, hazard increases over time

---

## Effect of location ($\mu$): shifting the survival curve

Holding $\sigma = 0.8$ fixed, vary $\mu$ from 0.5 to 3.0.

```{r location-effect}
#| code-fold: true
weibull_S <- function(t, mu, sigma) {
  exp(-exp((log(t) - mu) / sigma))
}

t_grid <- seq(0.01, 30, length.out = 300)
mu_vals <- c(0.5, 1.0, 1.5, 2.0, 2.5, 3.0)
sigma_fixed <- 0.8

df <- do.call(rbind, lapply(mu_vals, function(mu) {
  data.frame(time = t_grid,
             survival = weibull_S(t_grid, mu, sigma_fixed),
             mu = paste0("mu = ", mu))
}))

g2(df, survival ~ time, color = "mu",
   title = "Weibull survival: varying location (sigma = 0.8)") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "S(t)")
```

Larger $\mu$ = longer expected survival time. The curves shift right.

---

## Location effect on the hazard function

```{r location-hazard}
#| code-fold: true
weibull_h <- function(t, mu, sigma) {
  alpha <- 1 / sigma
  lambda <- exp(-mu / sigma)
  alpha * lambda * t^(alpha - 1)
}

df_h <- do.call(rbind, lapply(c(1.0, 2.0, 3.0), function(mu) {
  data.frame(time = t_grid,
             hazard = weibull_h(t_grid, mu, sigma_fixed),
             mu = paste0("mu = ", mu))
}))

g2(df_h, hazard ~ time, color = "mu",
   title = "Weibull hazard: varying location (sigma = 0.8)") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "h(t)")
```

The hazard curves shift right and downward with increasing $\mu$.
Under constant $\sigma$, the **hazard ratio is constant** — proportional hazards.

---

## Effect of scale ($\sigma$): reshaping the curve

Holding $\mu = 2.0$ fixed, vary $\sigma$ from 0.3 to 1.5.

```{r scale-effect}
#| code-fold: true
sigma_vals <- c(0.3, 0.5, 0.8, 1.0, 1.5)

df_s <- do.call(rbind, lapply(sigma_vals, function(s) {
  data.frame(time = t_grid,
             survival = weibull_S(t_grid, 2.0, s),
             sigma = paste0("sigma = ", s))
}))

g2(df_s, survival ~ time, color = "sigma",
   title = "Weibull survival: varying scale (mu = 2.0)") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "S(t)")
```

- Small $\sigma$ (large $\alpha = 1/\sigma$): steep drop, concentrated risk
- Large $\sigma$: gradual decline, drawn-out tail

---

## Scale effect on the hazard

```{r scale-hazard}
#| code-fold: true
df_sh <- do.call(rbind, lapply(c(0.5, 0.8, 1.0, 1.5), function(s) {
  h <- weibull_h(t_grid, 2.0, s)
  h[h > 2] <- NA  # cap for display
  data.frame(time = t_grid, hazard = h,
             sigma = paste0("sigma = ", s))
}))
df_sh <- df_sh[!is.na(df_sh$hazard), ]

g2(df_sh, hazard ~ time, color = "sigma",
   title = "Weibull hazard: varying scale (mu = 2.0)") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "h(t)")
```

- $\sigma < 1$ ($\alpha > 1$): **increasing** hazard — aging / wear-out
- $\sigma = 1$ ($\alpha = 1$): **constant** hazard — exponential
- $\sigma > 1$ ($\alpha < 1$): **decreasing** hazard — early risk

---

## Combining location and scale

```{r location-scale-combined}
#| code-fold: true
combos <- list(
  list(mu = 1.0, sigma = 0.5, label = "mu=1.0, sigma=0.5 (high risk, steep)"),
  list(mu = 2.0, sigma = 0.5, label = "mu=2.0, sigma=0.5 (low risk, steep)"),
  list(mu = 1.0, sigma = 1.2, label = "mu=1.0, sigma=1.2 (high risk, gradual)"),
  list(mu = 2.0, sigma = 1.2, label = "mu=2.0, sigma=1.2 (low risk, gradual)")
)

df_c <- do.call(rbind, lapply(combos, function(p) {
  data.frame(time = t_grid,
             survival = weibull_S(t_grid, p$mu, p$sigma),
             profile = p$label)
}))

g2(df_c, survival ~ time, color = "profile",
   title = "Weibull survival: four location-scale combinations") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "S(t)")
```

Location shifts the curve; scale controls how abruptly risk accumulates.

---

## Why constant $\sigma$ is restrictive

Under the standard Weibull AFT with constant $\sigma$:

$$
\frac{h_A(t)}{h_B(t)} = \text{constant for all } t
$$

This is **proportional hazards**. In practice, hazard ratios often:

- Converge toward 1 over time (treatment effect wanes)
- Diverge (delayed treatment effect)
- Cross (early harm, late benefit)

A single extra parameter can capture these patterns.

---

## The VLDAFT extension (Anderson, 1991)

Let $\sigma$ vary with the fitted location:

$$
\log \sigma_i = \gamma_0 + \theta_1 (\mu_i - \bar\mu)
$$

- $\theta_1 = 0$: standard AFT (proportional hazards under Weibull)
- $\theta_1 < 0$: subjects with longer expected survival have **tighter** dispersion → **converging hazards**
- $\theta_1 > 0$: **diverging hazards**

One scalar parameter breaks proportional hazards.

---

## Visualising the $\theta_1$ effect

```{r theta-effect}
#| code-fold: true
mu_low <- 1.5; mu_high <- 2.5; mubar <- 2.0; gamma0 <- log(0.6)
theta_vals <- c(0, -0.3, -0.6, 0.3)

df_theta <- do.call(rbind, lapply(theta_vals, function(th) {
  sig_low  <- exp(gamma0 + th * (mu_low  - mubar))
  sig_high <- exp(gamma0 + th * (mu_high - mubar))
  rbind(
    data.frame(time = t_grid,
               survival = weibull_S(t_grid, mu_low, sig_low),
               group = "high risk (mu=1.5)",
               theta = paste0("theta1 = ", th)),
    data.frame(time = t_grid,
               survival = weibull_S(t_grid, mu_high, sig_high),
               group = "low risk (mu=2.5)",
               theta = paste0("theta1 = ", th))
  )
}))

g2(df_theta[df_theta$theta == "theta1 = 0", ],
   survival ~ time, color = "group",
   title = "theta1 = 0: proportional hazards") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "S(t)")
```

---

## Converging hazards: $\theta_1 = -0.3$

```{r theta-neg}
#| code-fold: true
g2(df_theta[df_theta$theta == "theta1 = -0.3", ],
   survival ~ time, color = "group",
   title = "theta1 = -0.3: converging hazards") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "S(t)")
```

The low-risk and high-risk curves are closer together at late times.
Hazards converge — the risk-factor effect **wanes** over time.

---

## Anderson (1991): Framingham Heart Study

- 1,433 men aged 30–49, followed ~32 years for CHD incidence
- Covariates: log(age), log(SBP), log(cholesterol), relative weight, smoking
- Standard Weibull AFT (Model 1) vs VLDAFT with $\theta_1$ (Model 2)

**Key finding:** $\hat\theta_1 = -0.95$ (highly significant, LRT $\chi^2_1 = 20.9$, $p < 0.001$)

Interpretation: the dispersion **increases** with predicted survival time,
meaning risk-factor effects on the hazard **converge** over the 32-year follow-up.

The standard proportional hazards model was rejected in favor of converging hazards.

---

## Anderson (1991): model comparison

| Model | Description | Parameters | Log-likelihood |
|-------|-------------|:----------:|:--------------:|
| 1 | Standard Weibull AFT | 7 | −4414.5 |
| 2 | VLDAFT ($\theta_1$) | 8 | −4404.0 |
| 3a | VLDAFT ($\theta_1, \theta_2$) | 9 | −4403.9 |
| 3b | Full scale regression | 12 | −4404.0 |

- Model 2 strongly beats Model 1 (LRT = 20.9, 1 df)
- Models 3a and 3b offer no improvement over Model 2
- **One parameter ($\theta_1$) captures the entire nonproportional hazards effect**

---

## Adding a cure fraction

Some diseases have long-term survivors who will never experience the event.
The **Poisson-mixture cure model** adds a plateau:

$$
S(t) = \exp\{-\tau(1 - S_0(t))\}, \qquad \tau = -\log(p)
$$

where $p$ is the cure probability and $S_0(t)$ is the baseline Weibull survival.

As $t \to \infty$: $S(t) \to \exp(-\tau) = p$.

---

## Cure rate effect on survival

Fix $\mu = 2.0$, $\sigma = 0.8$, vary cure rate $p$.

```{r cure-rate-effect}
#| code-fold: true
cure_S <- function(t, mu, sigma, p) {
  S0 <- weibull_S(t, mu, sigma)
  tau <- -log(p)
  exp(-tau * (1 - S0))
}

cure_rates <- c(0, 0.1, 0.3, 0.5)
df_cure <- do.call(rbind, lapply(cure_rates, function(p) {
  S <- if (p == 0) weibull_S(t_grid, 2.0, 0.8)
       else cure_S(t_grid, 2.0, 0.8, p)
  data.frame(time = t_grid, survival = S,
             cure = paste0("p = ", sprintf("%.0f%%", p * 100)))
}))

g2(df_cure, survival ~ time, color = "cure",
   title = "Effect of cure rate on Weibull survival (mu=2, sigma=0.8)") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "S(t)")
```

The plateau at $S(\infty) = p$ is clearly visible for $p > 0$.

---

## Cure model: shape parameter effect

Fix $\mu = \log(2) \approx 0.69$ (median ~2 years for uncured), cure rate $p = 0.3$. Vary $\sigma$.

```{r cure-shape}
#| code-fold: true
mu_fixed <- log(2)
p_fixed <- 0.3
sigma_cure_vals <- c(0.3, 0.5, 0.8, 1.0, 1.5)

t_grid5 <- seq(0.01, 5, length.out = 300)

df_cs <- do.call(rbind, lapply(sigma_cure_vals, function(s) {
  data.frame(
    time = t_grid5,
    survival = cure_S(t_grid5, mu_fixed, s, p_fixed),
    sigma = paste0("sigma = ", s, " (alpha = ", round(1/s, 2), ")")
  )
}))

g2(df_cs, survival ~ time, color = "sigma",
   title = "Cure Weibull: varying shape (mu=log(2), cure=30%)") |>
  mark_line() |>
  scale_x(domain = c(0, 5), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time (years)") |>
  axis_y(title = "S(t)")
```

- Small $\sigma$: sharp early decline, then quick approach to the 30% plateau
- Large $\sigma$: gradual decline, slow approach to the plateau

All curves converge to $S(\infty) = 0.30$.

---

## Summary

| Feature | What it controls | Key insight |
|---------|-----------------|-------------|
| $\mu$ (location) | Shifts survival curve in time | Higher $\mu$ = longer expected survival |
| $\sigma$ (scale) | Shape of hazard: increasing, constant, decreasing | $\sigma < 1$: wear-out; $\sigma > 1$: early risk |
| $\theta_1$ (coupling) | Non-proportional hazards | One parameter breaks PH assumption |
| $p$ (cure rate) | Long-run survival plateau | $S(\infty) = p$ for cured fraction |

The `vldaft` package implements all of these in a unified framework with C and Rust backends, `gamlss2` integration, and support for right/left/interval censoring and left truncation.

---

## References

- Anderson, K. M. (1991). A nonproportional hazards Weibull accelerated
  failure time regression model. *Biometrics*, **47**(1), 281–288.
- Rodrigues, J., Cancho, V. G., de Castro, M. & Louzada-Neto, F.
  (2009). On the unification of long-term survival models.
  *Statistics & Probability Letters*, **79**(6), 753–759.
- Package source: [github.com/keaven/vldaft](https://github.com/keaven/vldaft)
