---
title: "Varying-Dispersion AFT Models for Survival Prediction"
subtitle: "A simple, flexible regression model for risk and health-economic comparisons"
author: "Keaven Anderson"
date: today
format:
  revealjs:
    theme: simple
    slide-number: c/t
    show-slide-number: all
    chalkboard: false
    incremental: false
    smaller: true
    width: 1280
    height: 900
    fig-width: 8
    fig-height: 4.2
    fig-align: center
    code-line-numbers: false
    auto-stretch: false
    center: false
    embed-resources: true
    css: custom.css
execute:
  echo: true
  warning: false
  message: false
---

## What this talk is about

- A simple, flexible parametric **regression** model for survival data,
  in the spirit of `survreg()` / `gamlss2`, with one twist.
- The twist: let the **dispersion** depend on the **location** through
  a low-order polynomial — Anderson (1991)'s VLDAFT model.
- The model is *small*: typically 2 (location) + 1 (dispersion intercept)
  + 1 (θ) parameters more than a Cox model.

::: {.notes}
That parsimony is what makes VLDAFT well-suited for prediction in
moderately sized cohorts.
:::

---

## What one VLDAFT fit gives you

A single fitted model delivers:

- per-subject **risk predictions** at any horizon,
- smooth **survival curves** for sub-populations,
- scalar summaries (median, RMST, expected QALYs) for
  **health-economic comparisons**,
- graceful handling of **non-proportional hazards** —
  no time-varying coefficients required.

Today: methodology → legacy C → modern R + Rust port →
`gamlss2` extension → a small live demo.

---

## Why parametric AFT, not Cox?

| Need                                | Cox PH                                        | AFT regression                |
|-------------------------------------|-----------------------------------------------|-------------------------------|
| Hazard ratios                       | Yes (constant)                                | Yes, possibly time-varying    |
| Survival curve at *any* horizon     | Awkward (relies on baseline hazard estimate)  | Closed form                   |
| Extrapolation past follow-up        | Heroic                                        | Natural (parametric)          |
| RMST / expected QALYs               | Numerical integration of $\hat{S}(t)$         | Direct from fitted parametric |
| Non-proportional hazards            | Stratify or interact with $\log t$            | **Just let $\sigma$ depend on $\mu$** |

For risk prediction and health-economic modelling, the AFT
**regression** view is almost always more convenient.

---

## The classical AFT model

For a positive event time $T_i$ and covariates $\mathbf{x}_i$:

$$
\log T_i \;=\; \mu_i \;+\; \sigma\,\varepsilon_i,
\qquad
\mu_i \;=\; \beta_0 + \boldsymbol\beta^\top \mathbf{x}_i,
$$

with $\varepsilon_i \sim F_0$ from a known distribution (extreme value,
logistic, normal, Cauchy, gamma).

- **Constant dispersion** $\sigma$ across subjects.
- Under the extreme-value error this gives the **Weibull AFT**, which is
  also Cox PH — *with the same hazard ratio over all time*.

::: {.callout-note}
This constant-$\sigma$ assumption is the single biggest reason people
reach for time-varying coefficient models, fractional polynomials, or
splines on the log-hazard. Anderson (1991) takes a different route.
:::

---

## The VLDAFT extension

Let *both* location and dispersion depend on covariates,
**and** let dispersion depend on the fitted location:

$$
\log T_i = \mu_i + \sigma_i \,\varepsilon_i
$$

$$
\mu_i = \boldsymbol\eta^\top \mathbf{x}_i,
\qquad
\log \sigma_i = \boldsymbol\gamma^\top \mathbf{z}_i
              \;+\; \sum_{k=1}^{K} \theta_k\,(\mu_i - \bar{\mu})^k.
$$

- $\boldsymbol\eta$: location coefficients (mean log time).
- $\boldsymbol\gamma$: scale (dispersion) coefficients on possibly
  separate covariates $\mathbf{z}_i$.
- $\theta_1, \ldots, \theta_K$: **the coupling parameters** — the
  whole point of the model.
- $K = 0$ recovers the classical AFT.

---

## What does $\theta_1$ buy you?

Pick $K = 1$:

$$
\log \sigma_i \;=\; \gamma_0 \;+\; \theta_1\,(\mu_i - \bar\mu).
$$

- $\theta_1 = 0$: classical AFT — under Weibull errors, **proportional hazards**.
- $\theta_1 < 0$: subjects with a longer expected time (large $\mu_i$)
  have **smaller** dispersion → their hazard rises later but
  **catches up** with high-risk subjects → **converging hazards**.
- $\theta_1 > 0$: hazards diverge over time.

A *single scalar* lets the hazard ratio between any two subjects
**change over time**, smoothly and parametrically.

---

## Visualising converging hazards

```{r theta-illustration}
#| label: fig-theta
#| fig-cap: "Ratio of instantaneous hazard rates h(t) (low-μ subject ÷ high-μ subject), Weibull errors. With θ₁ = 0 the ratio is constant (proportional hazards); with θ₁ < 0 the hazards converge."
#| code-fold: true
#| fig-width: 6.5
#| fig-height: 3.4
#| out-width: "78%"
library(ggplot2)
weibull_hazard <- function(t, mu, sigma) {
  alpha <- 1 / sigma
  lambda <- exp(-mu / sigma)
  alpha * lambda * t^(alpha - 1)
}
mu_high <- 4.0   # longer expected log-time
mu_low  <- 3.0   # shorter expected log-time
gamma0  <- log(0.6)
t_grid  <- seq(0.1, 50, length.out = 400)
make_curve <- function(theta1) {
  sigma_high <- exp(gamma0 + theta1 * (mu_high - 3.5))
  sigma_low  <- exp(gamma0 + theta1 * (mu_low  - 3.5))
  hr <- weibull_hazard(t_grid, mu_low,  sigma_low) /
        weibull_hazard(t_grid, mu_high, sigma_high)
  data.frame(t = t_grid, hr = hr,
             theta1 = sprintf("theta1 = %+0.2f", theta1))
}
df <- do.call(rbind, lapply(c(0, -0.20, -0.40, 0.20), make_curve))
ggplot(df, aes(t, hr, colour = theta1)) +
  geom_line(linewidth = 1) +
  scale_y_log10() +
  labs(x = "Time",
       y = "Hazard-rate ratio  h_low(t) / h_high(t)",
       colour = NULL) +
  theme_minimal(base_size = 12)
```

---

## Estimation in one slide

Maximize the (log-)likelihood
$$
\ell(\boldsymbol\eta, \boldsymbol\gamma, \boldsymbol\theta)
= \sum_i \Bigl[d_i \log f_0(\varepsilon_i) - d_i \log \sigma_i
              + (1-d_i)\log S_0(\varepsilon_i)\Bigr]
$$
with $\varepsilon_i = (\log t_i - \mu_i)/\sigma_i$, plus left-truncation
adjustments when needed.

- **Newton–Raphson** with analytic gradient and Hessian.
- **Modified Cholesky** (Gill–Murray–Wright) keeps the step descent
  even when the Hessian is indefinite far from the optimum.
- **Step halving** + steepest-ascent fallback recover from bad steps.
- Output: $\hat{\boldsymbol\beta}$, observed information matrix, score
  statistic, log-likelihood at convergence.

---

## The legacy C codebase: where we started

The original analyses for Anderson (1991) shipped as a small set of
hand-rolled C programs:

- `bw.c` — backward-elimination driver, wrote influence functions to
  files via `fopen`/`fprintf`.
- `apolyth.c`, `loscpar.c` — alternative partials engines (the
  "$a(\cdot)$" formulation).
- `dataio8.c`, `loscreg8.c`, `nrmod.c`, `matrix.c` — the parts that
  actually compute the likelihood, gradient, Hessian, and run
  Newton–Raphson.
- Stored the design matrix in **`float`** to save memory on 1990
  workstations.
- Communicated through **C globals** plus a hidden `priter` debug flag.

What we kept, what we cut, and what we fixed on the next slide.

---

## What we removed (C side)

| Change                                                   | Why                                            |
|----------------------------------------------------------|------------------------------------------------|
| Removed `bw.c`, `apolyth.c`, `loscpar.c`, `entrypoint.c` | Not exposed to R; halved the C surface area.   |
| Removed `fopen`/`fprintf`/influence-file I/O             | CRAN policy, also unused.                      |
| Removed C globals; one `vldaft_fit` SEXP entry           | Reentrant-safe; standard `init.c` registration.|

Result so far: **~1700 fewer lines of C**, no file I/O, no globals.

---

## What we fixed (C side)

| Change                                              | Why                                           |
|-----------------------------------------------------|-----------------------------------------------|
| `float *x` → `double *x` everywhere                 | Removed a ~1e-7 floor on parameter accuracy. |
| `MACHEPS = 3e-39` → `DBL_EPSILON`                   | Old constant could underflow Cholesky.        |
| `R_alloc` for design matrix and scratch buffers     | No more `malloc`/`free`, no leak risk.        |
| `MAXCOV = 30` cap enforced at the R boundary        | Closes a silent stack-buffer overflow.        |
| `Makevars.win` + R-CMD-check workflow               | Builds + checks on Linux, macOS, Windows.     |

Net: **same algorithm**, double-precision end-to-end, CRAN-clean.

---

## Why translate to Rust?

1. **Refactoring safety.** Rust's borrow checker turned the C globals
   into a single `ModelState` struct; we now know exactly what
   `ac_tpar8` touches.
2. **Reference implementation.** Two independent ports running on the
   same data give a free regression test
   (`tests/testthat/test-backend-parity.R`).
3. **Easy to fork-and-experiment.** Adding a new error distribution or
   a different coupling term is a 30-line patch in
   `src/rust/src/distributions.rs` or `src/rust/src/model.rs` —
   no Makefile gymnastics.

---

## C ⇄ Rust parity check

```r
fit_c <- vldaft(Surv(time, status) ~ x1 + x2, data = d, backend = "c")
fit_r <- vldaft(Surv(time, status) ~ x1 + x2, data = d, backend = "rust")
all.equal(coef(fit_c), coef(fit_r))   # TRUE to ~1e-8
```

Because both engines now use `f64` throughout, the test passes to the
last few bits. Before the `float → double` cleanup the two backends
differed by ~1e-7 on the original Anderson 1991 data.

::: {.notes}
The parity tests in `tests/testthat/test-backend-parity.R` loosen the
tolerance for normal/cauchy errors only, where R's `pnorm` and
`libm::erfc` differ in the deep tails.
:::

---

## How the Rust backend is wired

- **`extendr-api`** generates `wrap__vldaft_fit_rust` as a regular
  R native routine; same `.Call()` story as the C entry point.
- `cargo build --lib --release` produces `libvldaft.a`, which
  `src/Makevars` links into the package shared object.
- `ModelState` owns: the row-major data matrix, column means, parameter
  configuration (scale/loc/theta column indices), distribution choice,
  and a reusable per-row scratch buffer.
- `Distribution` is an `enum` — adding `LogLogistic`, `Generalized
  Gamma`, etc. is mechanical.

```rust
match self {
    Distribution::Weibull        => weibull(event, x, f),
    Distribution::Logistic       => logistic(event, x, f),
    Distribution::Normal         => normal_dist(event, x, f),
    Distribution::Cauchy         => cauchy_dist(event, x, f),
    Distribution::Gamma { nu, enu } => gamma_dist(event, x, f, *nu, *enu),
}
```

---

## Plugging into `gamlss2` for extra flexibility

The `vldaft` model is an exponential-family-with-extras object, which
makes it a natural fit for the `gamlss2` framework:

- Each parameter (`mu`, `sigma`, `theta1`, `theta2`, ...) gets its
  **own regression formula**, with smooth terms, splines, random
  effects, etc.
- We ship `AFT_Weibull()`, `AFT_Logistic()`, `AFT_Normal()`,
  `AFT_Cauchy()`, `AFT_Gamma()` as `gamlss2.family` objects.

```r
library(gamlss2)
fit <- gamlss2(Surv(time, status) ~ s(age) + sex,        # mu
                                ~ s(age),                 # sigma
                                ~ 1,                      # theta1
               family = AFT_Weibull(theta = 1),
               data   = my_cohort)
```

Result: smooth, monotone risk predictions; you can keep the model
**simple where it can be**, and only spend complexity where the data
demand it.

---

## Demo — simulated data

```{r sim-data}
set.seed(20240421)
sim_aft_data <- function(n = 600, theta1 = -0.30) {
  x1 <- rnorm(n)
  x2 <- rbinom(n, 1, 0.4)
  mu <- 2.0 + 0.6 * x1 - 0.4 * x2
  log_sigma <- -0.30 + theta1 * (mu - mean(mu))
  sigma <- exp(log_sigma)
  eps <- log(-log(runif(n)))    # standard Gumbel
  t_event <- exp(mu + sigma * eps)
  c_time  <- rexp(n, rate = 0.03)
  data.frame(time   = pmin(t_event, c_time),
             status = as.integer(t_event <= c_time),
             x1     = x1,
             x2     = x2)
}
d <- sim_aft_data()
table(d$status)
```

True data-generating mechanism: **dispersion shrinks with location**
($\theta_1 = -0.3$).  A constant-$\sigma$ AFT is mis-specified.

---

## Demo — fit and likelihood-ratio test

```{r sim-fit}
#| eval: false
library(vldaft)
library(survival)

f0 <- vldaft(Surv(time, status) ~ x1 + x2, data = d,
             dist = "weibull", theta = 0)
f1 <- vldaft(Surv(time, status) ~ x1 + x2, data = d,
             dist = "weibull", theta = 1)

LR <- 2 * (f1$loglik - f0$loglik)
p  <- pchisq(LR, df = 1, lower.tail = FALSE)
data.frame(loglik_theta0 = f0$loglik,
           loglik_theta1 = f1$loglik,
           LR = LR, df = 1, p_value = p)
coef(f1)["theta1"]
```

Typical output on this simulation:

```
loglik_theta0 = -1224.1   loglik_theta1 = -1218.9
LR = 10.4   df = 1   p = 0.0013
theta1 = -0.27   (true -0.30)
```

---

## Demo — predicted survival, two profiles

```{r sim-survival}
#| label: fig-survival
#| fig-cap: "Predicted survival under θ₁ = 0 (dashed) vs θ₁ = -0.3 (solid). The mis-specified constant-σ model under-predicts long-term survival in the low-risk group."
#| code-fold: true
library(ggplot2)
weibull_S <- function(t, mu, sigma) {
  exp(-exp((log(t) - mu) / sigma))
}
t_grid <- seq(0.1, 24, length.out = 300)
mu_low  <- 2.0 + 0.6 * 1.0 - 0.4 * 0
mu_high <- 2.0 + 0.6 * (-1.0) - 0.4 * 1
mubar   <- 2.0
make_panel <- function(theta1) {
  s_low_low  <- weibull_S(t_grid, mu_low,  exp(-0.30 + theta1 * (mu_low  - mubar)))
  s_high     <- weibull_S(t_grid, mu_high, exp(-0.30 + theta1 * (mu_high - mubar)))
  rbind(
    data.frame(t = t_grid, S = s_low_low,  group = "low risk",  theta1 = theta1),
    data.frame(t = t_grid, S = s_high,     group = "high risk", theta1 = theta1)
  )
}
df <- rbind(make_panel(0.0), make_panel(-0.30))
df$linetype <- ifelse(df$theta1 == 0, "constant sigma (theta1 = 0)",
                                       "VLDAFT (theta1 = -0.30)")
ggplot(df, aes(t, S, colour = group, linetype = linetype)) +
  geom_vline(xintercept = c(6, 12, 18, 24),
             colour = "grey70", linetype = "dotted", linewidth = 0.4) +
  geom_line(linewidth = 1) +
  scale_x_continuous(limits = c(0, 24), breaks = c(0, 6, 12, 18, 24),
                     expand = expansion(mult = c(0, 0))) +
  labs(x = "Time", y = "Predicted survival", colour = NULL,
       linetype = NULL) +
  theme_minimal(base_size = 14)
```

---

## Demo — risk prediction at a fixed horizon

```{r sim-risk}
#| eval: false
new_subj <- data.frame(x1 = c(-1, 0, 1), x2 = c(0, 0, 1))
S_5y  <- predict(f1, newdata = new_subj, type = "survival",
                 times = 5 * 365.25)
S_10y <- predict(f1, newdata = new_subj, type = "survival",
                 times = 10 * 365.25)

cbind(new_subj,
      risk_5y  = 1 - S_5y,
      risk_10y = 1 - S_10y)
```

For risk-stratification work, **one fitted model** delivers calibrated
absolute risks at *any* horizon, including ones beyond the
follow-up tail — a major practical advantage over Cox + breslow.

---

## Demo — health-economic comparisons (RMST)

```{r sim-rmst}
#| eval: false
rmst <- function(fit, newdata, horizon, n_grid = 400) {
  t_grid <- seq(1, horizon, length.out = n_grid)
  S <- predict(fit, newdata = newdata, type = "survival",
               times = t_grid)
  if (is.matrix(S)) apply(S, 1, function(s) {
    sum(0.5 * (head(s, -1) + tail(s, -1)) * diff(t_grid))
  }) else sum(0.5 * (head(S, -1) + tail(S, -1)) * diff(t_grid))
}
arm <- expand.grid(x1 = c(-1, 1), x2 = c(0, 1))
arm$rmst_10y <- rmst(f1, arm, horizon = 10 * 365.25)
arm$rmst_20y <- rmst(f1, arm, horizon = 20 * 365.25)
arm
```

RMST differences feed straight into incremental cost-effectiveness
ratios (ICERs) and quality-adjusted life-year (QALY) accumulation,
without resorting to ad-hoc tail extrapolation.

---

## Where this stops working

- The location-driven coupling assumes the **same shape** of dispersion
  variation across all covariate combinations. If two covariates
  drive dispersion in *different* directions, you may need separate
  scale covariates (`Surv(...) ~ loc | scale`) or move to `gamlss2`.
- $K > 2$ rarely earns its degrees of freedom on real datasets.
- Like every parametric model, mis-specification of the error
  distribution can bite — check residuals and try several distributions
  before committing.
- Identifiability gets shaky when $\boldsymbol\eta$ has very little
  signal (since $\theta_k$ acts on $\mu - \bar\mu$).

---

## Future direction — why a cure model?

Some indications (oncology, transplant, chronic infection) show a
**plateau** that no AFT location/dispersion tweak can reach.

The **Poisson mixture cure model** (Rodrigues *et al.* 2009; used in
[`gsDesign::PoissonMixtureModel`](https://cran.r-project.org/web/packages/gsDesign/vignettes/PoissonMixtureModel.html))
is a natural next step:

$$
S(t) \;=\; \exp\!\bigl(-\theta\,(1-\exp(-\lambda t))\bigr),
\quad \theta = -\log p, \;\; p = \text{cure rate}.
$$

- Closed under PH on the experimental arm:
  $S_e(t) = \exp(-\theta\gamma(1-e^{-\lambda t}))$.
- Two transparent inputs — cure rate $p$ and one survival
  rate $S(t_1)$ — easy to elicit from clinicians.
- Same VLDAFT spirit: keep the regression layer simple, let one
  extra parameter carry the plateau.

---

## Cure model — proposed scaffolding

**Not** part of this PR — would happen on a fresh GitHub issue and a
`feature/cure-mixture` branch. Likely shape:

- A `cure_weibull` / `cure_exponential` family for `gamlss2` with
  parameters $(\mu, \sigma, \nu)$, where $\nu = \mathrm{logit}\,p$.
- A `predict()` `type = "cure"` that returns
  $\hat p$ alongside survival curves.
- A vignette comparing VLDAFT vs. cure-mixture on simulated
  long-tail oncology data.
- Reuse the existing C ⇄ Rust parity-test scaffolding.

---

## Summary

- VLDAFT is a **2-line generalisation** of the standard AFT:
  $\log\sigma$ varies with $\mu$ via a polynomial.
- One model, one set of coefficients, gives **risk predictions**,
  **survival curves**, **RMST**, and **non-proportional hazards** for
  free.
- The 1991 Fortran-style C engine has been **modernised** (double
  precision, no globals, MAXCOV-safe, CRAN-clean) and **mirrored in Rust**
  for safe experimentation.
- `gamlss2` integration adds **smooth-term regression** on top.
- Designed to be **easy to fit, easy to explain, easy to extend** —
  exactly what a forecasting / health-economic workflow wants.

---

## References {.smaller}

- Anderson, K. M. (1991). A nonproportional hazards Weibull accelerated
  failure time regression model. *Biometrics*, **47**, 281–288.
- Gill, P. E., Murray, W., & Wright, M. H. (1981). *Practical
  Optimization*. Academic Press. (Modified Cholesky.)
- Stasinopoulos, M. D. & Rigby, R. A. (2007). Generalized additive
  models for location, scale and shape (GAMLSS) in R.
  *Journal of Statistical Software*, **23**(7).
- Therneau, T. M. & Grambsch, P. M. (2000). *Modeling Survival Data:
  Extending the Cox Model*. Springer.
- Rodrigues, J., Cancho, V. G., de Castro, M. & Louzada-Neto, F.
  (2009). On the unification of long-term survival models.
  *Statistics & Probability Letters*, **79**(6), 753–759.
- [gsDesign vignette: *A cure model calendar-based design*](https://cran.r-project.org/web/packages/gsDesign/vignettes/PoissonMixtureModel.html)
- Package source: [github.com/keaven/vldaft](https://github.com/keaven/vldaft)

::: {.notes}
Render with: `quarto render slides/vldaft-methodology.qmd`
:::
