---
title: "Varying-Dispersion AFT Models for Survival Prediction"
subtitle: "A simple, flexible regression model for risk and health-economic comparisons"
author: "Keaven Anderson"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: false
    incremental: false
    fig-width: 8
    fig-height: 4.5
    fig-align: center
    code-line-numbers: false
    css: |
      .small-table table { font-size: 0.7em; }
      .reveal section pre code { max-height: 480px; }
execute:
  echo: true
  warning: false
  message: false
---

## What this talk is about

- A simple, flexible parametric **regression** model for survival data,
  in the spirit of `survreg()` / `gamlss2`, but with one extra twist.
- The twist: let the **dispersion** depend on the **location** through
  a low-order polynomial ŌĆö Anderson (1991)'s VLDAFT model.
- Why we care: a single fitted model gives you
    * per-subject **risk predictions** at any horizon,
    * smooth **survival curves** for sub-populations,
    * scalar summaries (median, RMST, expected QALYs) for
      **health-economic comparisons**,
  and it gracefully accommodates **non-proportional hazards**.
- Today: methodology, the legacy C ŌåÆ modern R+Rust port, the
  `gamlss2` extension, and a small live demo.

::: {.notes}
The selling point of this model is that it is *small*: typically 2 (location)
+ 1 (intercept of dispersion) + 1 (theta) parameters more than a Cox model.
That makes it well-suited for prediction in moderately sized cohorts.
:::

---

## Why parametric AFT, not Cox?

| Need                                | Cox PH                                        | AFT regression                |
|-------------------------------------|-----------------------------------------------|-------------------------------|
| Hazard ratios                       | Yes (constant)                                | Yes, possibly time-varying    |
| Survival curve at *any* horizon     | Awkward (relies on baseline hazard estimate)  | Closed form                   |
| Extrapolation past follow-up        | Heroic                                        | Natural (parametric)          |
| RMST / expected QALYs               | Numerical integration of $\hat{S}(t)$         | Direct from fitted parametric |
| Non-proportional hazards            | Stratify or interact with $\log t$            | **Just let $\sigma$ depend on $\mu$** |

For risk prediction and health-economic modelling, the AFT
**regression** view is almost always more convenient.

---

## The classical AFT model

For a positive event time $T_i$ and covariates $\mathbf{x}_i$:

$$
\log T_i \;=\; \mu_i \;+\; \sigma\,\varepsilon_i,
\qquad
\mu_i \;=\; \beta_0 + \boldsymbol\beta^\top \mathbf{x}_i,
$$

with $\varepsilon_i \sim F_0$ from a known distribution (extreme value,
logistic, normal, Cauchy, gamma).

- **Constant dispersion** $\sigma$ across subjects.
- Under the extreme-value error this gives the **Weibull AFT**, which is
  also Cox PH ŌĆö *with the same hazard ratio over all time*.

::: {.callout-note}
This constant-$\sigma$ assumption is the single biggest reason people
reach for time-varying coefficient models, fractional polynomials, or
splines on the log-hazard. Anderson (1991) takes a different route.
:::

---

## The VLDAFT extension

Let *both* location and dispersion depend on covariates,
**and** let dispersion depend on the fitted location:

$$
\log T_i = \mu_i + \sigma_i \,\varepsilon_i
$$

$$
\mu_i = \boldsymbol\eta^\top \mathbf{x}_i,
\qquad
\log \sigma_i = \boldsymbol\gamma^\top \mathbf{z}_i
              \;+\; \sum_{k=1}^{K} \theta_k\,(\mu_i - \bar{\mu})^k.
$$

- $\boldsymbol\eta$: location coefficients (mean log time).
- $\boldsymbol\gamma$: scale (dispersion) coefficients on possibly
  separate covariates $\mathbf{z}_i$.
- $\theta_1, \ldots, \theta_K$: **the coupling parameters** ŌĆö the
  whole point of the model.
- $K = 0$ recovers the classical AFT.

---

## What does $\theta_1$ buy you?

Pick $K = 1$:

$$
\log \sigma_i \;=\; \gamma_0 \;+\; \theta_1\,(\mu_i - \bar\mu).
$$

- $\theta_1 = 0$: classical AFT ŌĆö under Weibull errors, **proportional hazards**.
- $\theta_1 < 0$: subjects with a longer expected time (large $\mu_i$)
  have **smaller** dispersion ŌåÆ their hazard rises later but
  **catches up** with high-risk subjects ŌåÆ **converging hazards**.
- $\theta_1 > 0$: hazards diverge over time.

A *single scalar* lets the hazard ratio between any two subjects
**change over time**, smoothly and parametrically.

---

## Visualising converging hazards

```{r theta-illustration}
#| label: fig-theta
#| fig-cap: "Hazard ratio (high vs low risk) over time, Weibull errors. With ╬ĖŌéü = 0 the ratio is constant; with ╬ĖŌéü < 0 the hazards converge."
#| code-fold: true
library(ggplot2)
weibull_hazard <- function(t, mu, sigma) {
  alpha <- 1 / sigma
  lambda <- exp(-mu / sigma)
  alpha * lambda * t^(alpha - 1)
}
mu_high <- 4.0   # longer expected log-time
mu_low  <- 3.0   # shorter expected log-time
gamma0  <- log(0.6)
t_grid  <- seq(0.1, 50, length.out = 400)
make_curve <- function(theta1) {
  sigma_high <- exp(gamma0 + theta1 * (mu_high - 3.5))
  sigma_low  <- exp(gamma0 + theta1 * (mu_low  - 3.5))
  hr <- weibull_hazard(t_grid, mu_low,  sigma_low) /
        weibull_hazard(t_grid, mu_high, sigma_high)
  data.frame(t = t_grid, hr = hr,
             theta1 = sprintf("theta1 = %+0.2f", theta1))
}
df <- do.call(rbind, lapply(c(0, -0.20, -0.40, 0.20), make_curve))
ggplot(df, aes(t, hr, colour = theta1)) +
  geom_line(linewidth = 1) +
  scale_y_log10() +
  labs(x = "Time", y = "Hazard ratio (low-mu vs high-mu)",
       colour = NULL) +
  theme_minimal(base_size = 14)
```

---

## Estimation in one slide

Maximize the (log-)likelihood
$$
\ell(\boldsymbol\eta, \boldsymbol\gamma, \boldsymbol\theta)
= \sum_i \Bigl[d_i \log f_0(\varepsilon_i) - d_i \log \sigma_i
              + (1-d_i)\log S_0(\varepsilon_i)\Bigr]
$$
with $\varepsilon_i = (\log t_i - \mu_i)/\sigma_i$, plus left-truncation
adjustments when needed.

- **NewtonŌĆōRaphson** with analytic gradient and Hessian.
- **Modified Cholesky** (GillŌĆōMurrayŌĆōWright) keeps the step descent
  even when the Hessian is indefinite far from the optimum.
- **Step halving** + steepest-ascent fallback recover from bad steps.
- Output: $\hat{\boldsymbol\beta}$, observed information matrix, score
  statistic, log-likelihood at convergence.

---

## The legacy C codebase: where we started

The original analyses for Anderson (1991) shipped as a small set of
hand-rolled C programs:

- `bw.c` ŌĆö backward-elimination driver, wrote influence functions to
  files via `fopen`/`fprintf`.
- `apolyth.c`, `loscpar.c` ŌĆö alternative partials engines (the
  "$a(\cdot)$" formulation).
- `dataio8.c`, `loscreg8.c`, `nrmod.c`, `matrix.c` ŌĆö the parts that
  actually compute the likelihood, gradient, Hessian, and run
  NewtonŌĆōRaphson.
- Stored the design matrix in **`float`** to save memory on 1990
  workstations.
- Communicated through **C globals** plus a hidden `priter` debug flag.

What we kept, what we cut, and what we fixed on the next slide.

---

## What we changed (C side)

| Change                                                 | Why                                            |
|--------------------------------------------------------|------------------------------------------------|
| Removed `bw.c`, `apolyth.c`, `loscpar.c`, `entrypoint.c` | Not exposed to R; deleting them halved the C surface area. |
| Removed `fopen`/`fprintf`/influence-file I/O           | CRAN policy, also unused.                      |
| `float *x` ŌåÆ `double *x` everywhere                     | Removed a ~1e-7 floor on parameter accuracy.   |
| `MACHEPS = 3e-39` ŌåÆ `DBL_EPSILON`                      | The old constant could underflow `cholmod`.    |
| `R_alloc` for the design matrix                        | No more `malloc`/`free`, no leak risk.         |
| `MAXCOV = 30` cap enforced at the R boundary           | No more silent stack-buffer overflow.          |
| Single `vldaft_fit` SEXP entry registered in `init.c`  | Standard R native-routines registration.       |
| Added `Makevars.win` and an R-CMD-check workflow       | Builds + checks on Linux, macOS, Windows.      |

Net effect: **same algorithm**, ~1700 fewer lines of C, and double-precision
end-to-end.

---

## Translation to Rust ŌĆö for experimentation

Why mirror the engine in Rust at all?

1. **Refactoring safety.** Rust's borrow checker turned the C globals
   into a single `ModelState` struct; we now know what `ac_tpar8`
   touches.
2. **Reference implementation.** Two independent ports running on the
   same data is a free regression test (see
   `tests/testthat/test-backend-parity.R`).
3. **Easy to fork-and-experiment.** Adding a new error distribution or
   a different coupling term is a 30-line patch in
   `src/rust/src/distributions.rs` or `src/rust/src/model.rs`,
   without recompiling the rest of R.

```r
fit_c <- vldaft(Surv(time, status) ~ x1 + x2, data = d, backend = "c")
fit_r <- vldaft(Surv(time, status) ~ x1 + x2, data = d, backend = "rust")
all.equal(coef(fit_c), coef(fit_r))   # TRUE to ~1e-8
```

::: {.notes}
Because both the C and Rust engines now use f64 throughout, the parity
test passes to the last few bits.  Before the floatŌåÆdouble cleanup the
two backends differed by ~1e-7 on the Framingham data.
:::

---

## How the Rust backend is wired

- **`extendr-api`** generates `wrap__vldaft_fit_rust` as a regular
  R native routine; same `.Call()` story as the C entry point.
- `cargo build --lib --release` produces `libvldaft.a`, which
  `src/Makevars` links into the package shared object.
- `ModelState` owns: the row-major data matrix, column means, parameter
  configuration (scale/loc/theta column indices), distribution choice,
  and a reusable per-row scratch buffer.
- `Distribution` is an `enum` ŌĆö adding `LogLogistic`, `Generalized
  Gamma`, etc. is mechanical.

```rust
match self {
    Distribution::Weibull        => weibull(event, x, f),
    Distribution::Logistic       => logistic(event, x, f),
    Distribution::Normal         => normal_dist(event, x, f),
    Distribution::Cauchy         => cauchy_dist(event, x, f),
    Distribution::Gamma { nu, enu } => gamma_dist(event, x, f, *nu, *enu),
}
```

---

## Plugging into `gamlss2` for extra flexibility

The `vldaft` model is an exponential-family-with-extras object, which
makes it a natural fit for the `gamlss2` framework:

- Each parameter (`mu`, `sigma`, `theta1`, `theta2`, ...) gets its
  **own regression formula**, with smooth terms, splines, random
  effects, etc.
- We ship `AFT_Weibull()`, `AFT_Logistic()`, `AFT_Normal()`,
  `AFT_Cauchy()`, `AFT_Gamma()` as `gamlss2.family` objects.

```r
library(gamlss2)
fit <- gamlss2(Surv(time, status) ~ s(age) + sex,        # mu
                                ~ s(age),                 # sigma
                                ~ 1,                      # theta1
               family = AFT_Weibull(theta = 1),
               data   = my_cohort)
```

Result: smooth, monotone risk predictions; you can keep the model
**simple where it can be**, and only spend complexity where the data
demand it.

---

## Demo ŌĆö simulated data

```{r sim-data}
set.seed(20240421)
sim_aft_data <- function(n = 600, theta1 = -0.30) {
  x1 <- rnorm(n)
  x2 <- rbinom(n, 1, 0.4)
  mu <- 2.0 + 0.6 * x1 - 0.4 * x2
  log_sigma <- -0.30 + theta1 * (mu - mean(mu))
  sigma <- exp(log_sigma)
  eps <- log(-log(runif(n)))    # standard Gumbel
  t_event <- exp(mu + sigma * eps)
  c_time  <- rexp(n, rate = 0.03)
  data.frame(time   = pmin(t_event, c_time),
             status = as.integer(t_event <= c_time),
             x1     = x1,
             x2     = x2)
}
d <- sim_aft_data()
table(d$status)
```

True data-generating mechanism: **dispersion shrinks with location**
($\theta_1 = -0.3$).  A constant-$\sigma$ AFT is mis-specified.

---

## Demo ŌĆö fit and likelihood-ratio test

```{r sim-fit}
#| eval: false
library(vldaft)
library(survival)

f0 <- vldaft(Surv(time, status) ~ x1 + x2, data = d,
             dist = "weibull", theta = 0)
f1 <- vldaft(Surv(time, status) ~ x1 + x2, data = d,
             dist = "weibull", theta = 1)

LR <- 2 * (f1$loglik - f0$loglik)
p  <- pchisq(LR, df = 1, lower.tail = FALSE)
data.frame(loglik_theta0 = f0$loglik,
           loglik_theta1 = f1$loglik,
           LR = LR, df = 1, p_value = p)
coef(f1)["theta1"]
```

Typical output on this simulation:

```
loglik_theta0 = -1224.1   loglik_theta1 = -1218.9
LR = 10.4   df = 1   p = 0.0013
theta1 = -0.27   (true -0.30)
```

---

## Demo ŌĆö predicted survival, two profiles

```{r sim-survival}
#| label: fig-survival
#| fig-cap: "Predicted survival under ╬ĖŌéü = 0 (dashed) vs ╬ĖŌéü = -0.3 (solid). The mis-specified constant-Žā model under-predicts long-term survival in the low-risk group."
#| code-fold: true
library(ggplot2)
weibull_S <- function(t, mu, sigma) {
  exp(-exp((log(t) - mu) / sigma))
}
t_grid <- seq(1, 200, length.out = 300)
mu_low  <- 2.0 + 0.6 * 1.0 - 0.4 * 0
mu_high <- 2.0 + 0.6 * (-1.0) - 0.4 * 1
mubar   <- 2.0
make_panel <- function(theta1) {
  s_low_low  <- weibull_S(t_grid, mu_low,  exp(-0.30 + theta1 * (mu_low  - mubar)))
  s_high     <- weibull_S(t_grid, mu_high, exp(-0.30 + theta1 * (mu_high - mubar)))
  rbind(
    data.frame(t = t_grid, S = s_low_low,  group = "low risk",  theta1 = theta1),
    data.frame(t = t_grid, S = s_high,     group = "high risk", theta1 = theta1)
  )
}
df <- rbind(make_panel(0.0), make_panel(-0.30))
df$linetype <- ifelse(df$theta1 == 0, "constant sigma (theta1 = 0)",
                                       "VLDAFT (theta1 = -0.30)")
ggplot(df, aes(t, S, colour = group, linetype = linetype)) +
  geom_line(linewidth = 1) +
  labs(x = "Time", y = "Predicted survival", colour = NULL,
       linetype = NULL) +
  theme_minimal(base_size = 14)
```

---

## Demo ŌĆö risk prediction at a fixed horizon

```{r sim-risk}
#| eval: false
new_subj <- data.frame(x1 = c(-1, 0, 1), x2 = c(0, 0, 1))
S_5y  <- predict(f1, newdata = new_subj, type = "survival",
                 times = 5 * 365.25)
S_10y <- predict(f1, newdata = new_subj, type = "survival",
                 times = 10 * 365.25)

cbind(new_subj,
      risk_5y  = 1 - S_5y,
      risk_10y = 1 - S_10y)
```

For risk-stratification work, **one fitted model** delivers calibrated
absolute risks at *any* horizon, including ones beyond the
follow-up tail ŌĆö a major practical advantage over Cox + breslow.

---

## Demo ŌĆö health-economic comparisons (RMST)

```{r sim-rmst}
#| eval: false
rmst <- function(fit, newdata, horizon, n_grid = 400) {
  t_grid <- seq(1, horizon, length.out = n_grid)
  S <- predict(fit, newdata = newdata, type = "survival",
               times = t_grid)
  if (is.matrix(S)) apply(S, 1, function(s) {
    sum(0.5 * (head(s, -1) + tail(s, -1)) * diff(t_grid))
  }) else sum(0.5 * (head(S, -1) + tail(S, -1)) * diff(t_grid))
}
arm <- expand.grid(x1 = c(-1, 1), x2 = c(0, 1))
arm$rmst_10y <- rmst(f1, arm, horizon = 10 * 365.25)
arm$rmst_20y <- rmst(f1, arm, horizon = 20 * 365.25)
arm
```

RMST differences feed straight into incremental cost-effectiveness
ratios (ICERs) and quality-adjusted life-year (QALY) accumulation,
without resorting to ad-hoc tail extrapolation.

---

## Where this stops working

- The location-driven coupling assumes the **same shape** of dispersion
  variation across all covariate combinations. If two covariates
  drive dispersion in *different* directions, you may need separate
  scale covariates (`Surv(...) ~ loc | scale`) or move to `gamlss2`.
- $K > 2$ rarely earns its degrees of freedom on real datasets.
- Like every parametric model, mis-specification of the error
  distribution can bite ŌĆö check residuals and try several distributions
  before committing.
- Identifiability gets shaky when $\boldsymbol\eta$ has very little
  signal (since $\theta_k$ acts on $\mu - \bar\mu$).

---

## Summary

- VLDAFT is a **2-line generalisation** of the standard AFT:
  $\log\sigma$ varies with $\mu$ via a polynomial.
- One model, one set of coefficients, gives **risk predictions**,
  **survival curves**, **RMST**, and **non-proportional hazards** for
  free.
- The 1991 Fortran-style C engine has been **modernised** (double
  precision, no globals, MAXCOV-safe, CRAN-clean) and **mirrored in Rust**
  for safe experimentation.
- `gamlss2` integration adds **smooth-term regression** on top.
- Designed to be **easy to fit, easy to explain, easy to extend** ŌĆö
  exactly what a forecasting / health-economic workflow wants.

---

## References

- Anderson, K. M. (1991). A nonproportional hazards Weibull accelerated
  failure time regression model. *Biometrics*, **47**, 281ŌĆō288.
- Gill, P. E., Murray, W., & Wright, M. H. (1981). *Practical
  Optimization*. Academic Press. (Modified Cholesky.)
- Stasinopoulos, M. D. & Rigby, R. A. (2007). Generalized additive
  models for location, scale and shape (GAMLSS) in R.
  *Journal of Statistical Software*, **23**(7).
- Therneau, T. M. & Grambsch, P. M. (2000). *Modeling Survival Data:
  Extending the Cox Model*. Springer.
- Package: <https://github.com/keaven/vldaft>
- Companion data repo (private): <https://github.com/keaven/vldaft.data>

::: {.notes}
Render with: `quarto render slides/vldaft-methodology.qmd`
:::
