---
title: "Reproducing Anderson (1991): A Nonproportional Hazards Weibull AFT Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing Anderson (1991): A Nonproportional Hazards Weibull AFT Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Load pre-computed summary results (sanitized; no raw data/model objects)
anderson_results <- readRDS(
  system.file("extdata", "anderson1991_results.rds", package = "vldaft"))
info <- anderson_results$info
summaries <- anderson_results$summaries
checks <- anderson_results$checks
prediction <- anderson_results$prediction
```

## Introduction

Anderson (1991) introduced the varying location and dispersion accelerated
failure time (VLDAFT) model, in which both the location $\mu$ and the
dispersion $\sigma$ of a log-linear survival model can depend on covariates
and on each other. This vignette reproduces the results from the original
Biometrics paper using the `vldaft` package.
This is done both with the original C routines to compute the log-likelihood and score functions, as well as through the `gamlss2` framework for comparison.
Both approaches yield the same results and replicate the paper.
We hope that this demonstration enables further use of the approach which can simplify modeling.
A particular advantage is on the Weibull distribution, where the VLDAFT model allows for nonproportional hazards by letting the dispersion depend on the location parameter.

**Reference:** Anderson, K.M. (1991). A nonproportional hazards Weibull
accelerated failure time regression model. *Biometrics*, **47**, 281--288.

## The Model

The VLDAFT model assumes

$$\Pr\!\left\{\frac{\log(T) - \mu}{\sigma} < u\right\} = F(u),$$

where $\mu = \boldsymbol{\beta}'\mathbf{X}$ is a linear function of
covariates, and the log-dispersion $\log(\sigma)$ can depend on both
covariates and on $\mu$ itself:

$$\log(\sigma) = \theta_0 + \theta_1 \mu^*.$$

Here $\mu^* = \mu - \bar\mu$ is the centered location parameter.
When $\theta_1 = 0$, this reduces to the standard linear location AFT
(LLAFT) model with proportional hazards under a Weibull distribution.

## Data

The example uses 1,433 men aged 30--49 from the Framingham Heart Study,
followed for approximately 32 years for coronary heart disease (CHD)
incidence. Risk factors were measured at baseline.

```{r data}
cat("Observations:", info$observations, "\n")
cat("CHD events:", info$chd_events, "\n")
cat("Censored:", info$censored, "\n")
```

The covariates are log(Age), log(systolic blood pressure),
log(cholesterol), Metropolitan relative weight, and a smoking indicator.

```{r summary}
print(info$covariate_summary)
```

## Model Fitting

Anderson (1991) reported four Weibull models of increasing complexity.
We reproduce all four.

### Model 1: Standard LLAFT (proportional hazards Weibull)

$$\mu = \boldsymbol{\beta}'\mathbf{X}, \quad \log(\sigma) = \theta_0$$

```{r model1, eval = FALSE}
fit1 <- vldaft(Surv(YTCHD, CHD) ~ ln.Age. + ln.SBP. + ln.Chol. + MRW + Smoke,
               data = MLT50, dist = "weibull", theta = 0)
```

```{r model1-summary}
cat(paste(summaries$fit1, collapse = "\n"), "\n")
```

Since Model 1 is a standard Weibull AFT, we can cross-check against
`survival::survreg`:

```{r survreg-check, eval = FALSE}
fit_survreg <- survreg(
  Surv(YTCHD, CHD) ~ ln.Age. + ln.SBP. + ln.Chol. + MRW + Smoke,
  data = MLT50, dist = "weibull")
```

```{r survreg-compare}
print(checks$coef_compare, digits = 7)
```

```{r scale-check}
# Compare scale parameter
cat("survreg scale (sigma):", checks$survreg_sigma, "\n")
cat("vldaft  scale (sigma):", checks$vldaft_sigma, "\n")
```

The coefficients and scale parameter agree, confirming the `vldaft`
implementation for the standard model.

### Model 2: VLDAFT with dispersion depending on location

$$\mu = \boldsymbol{\beta}'\mathbf{X}, \quad \log(\sigma) = \theta_0 + \theta_1 \mu^*$$

```{r model2, eval = FALSE}
fit2 <- vldaft(Surv(YTCHD, CHD) ~ ln.Age. + ln.SBP. + ln.Chol. + MRW + Smoke,
               data = MLT50, dist = "weibull", theta = 1)
```

```{r model2-summary}
cat(paste(summaries$fit2, collapse = "\n"), "\n")
```

### Model 3a: Quadratic theta

$$\mu = \boldsymbol{\beta}'\mathbf{X}, \quad \log(\sigma) = \theta_0 + \theta_1 \mu^* + \theta_2 {\mu^*}^2$$

As noted in the paper, convergence from zero starting values can be
difficult for more complex models. We use Model 2 estimates as
starting values, appending zero for the new $\theta_2$ parameter.

```{r model3a, eval = FALSE}
init3a <- c(fit2$coefficients, 0)
fit3a <- vldaft(Surv(YTCHD, CHD) ~ ln.Age. + ln.SBP. + ln.Chol. + MRW + Smoke,
                data = MLT50, dist = "weibull", theta = 2, init = init3a)
```

```{r model3a-summary}
cat(paste(summaries$fit3a, collapse = "\n"), "\n")
```

### Model 3b: Full scale model

$$\mu = \boldsymbol{\beta}'\mathbf{X}, \quad \log(\sigma) = \boldsymbol{\gamma}'\mathbf{X}$$

Here both location and scale have separate linear predictors with the
full set of covariates. The `|` operator in the formula separates
location (left) from scale (right) covariates.

```{r model3b, eval = FALSE}
fit3b <- vldaft(
  Surv(YTCHD, CHD) ~ ln.Age. + ln.SBP. + ln.Chol. + MRW + Smoke |
                      ln.Age. + ln.SBP. + ln.Chol. + MRW + Smoke,
  data = MLT50, dist = "weibull", theta = 0)
```

```{r model3b-summary}
cat(paste(summaries$fit3b, collapse = "\n"), "\n")
```

## Comparison with Published Results

The paper reports log-likelihoods on the time scale, which includes a
Jacobian term $\sum d_i \log(t_i)$. We adjust our log-likelihoods
accordingly.

```{r comparison}
print(checks$results_table, row.names = FALSE)
```

### Likelihood Ratio Test: Model 2 vs Model 1

```{r lrt}
cat(sprintf("LRT statistic: %.2f  (paper: 20.92)\n", checks$lrt))
cat(sprintf("p-value (chi-sq, 1 df): %.2e\n", checks$lrt_p))
```

The LLAFT model is strongly rejected in favor of the VLDAFT model
($\chi^2_1 = 20.9$, $p < 0.001$).

### Model 2 Coefficient Comparison

```{r coefs}
print(checks$coef_table, row.names = FALSE)
```

All coefficients match the published values to their reported precision.

## Fitting with gamlss2

The `vldaft` package provides `gamlss2`-compatible family objects
(`AFT_Weibull`, `AFT_Logistic`, etc.) so that the same models can be
fitted through the `gamlss2` framework. This enables the use of smooth
terms, additional diagnostics, and the full `gamlss2` infrastructure.

We demonstrate by fitting Model 2 (the VLDAFT model with
$\theta_1$ coupling) via `gamlss2`. In the `gamlss2` formula, the
`|` separator delineates submodels for each parameter: `mu | sigma | theta1`.

```{r gamlss2-model2, eval = FALSE}
library(gamlss2)

g2 <- gamlss2(
  Surv(YTCHD, CHD) ~ ln.Age. + ln.SBP. + ln.Chol. + MRW + Smoke | 1 | 1,
  data = MLT50, family = AFT_Weibull(theta = 1),
  control = gamlss2_control(maxit = 200))
```

```{r gamlss2-model2-summary, eval = FALSE}
summary(g2)
```

The `gamlss2` log-likelihood matches the `vldaft()` result:

```{r gamlss2-compare, eval = checks$has_gamlss2}
cat(sprintf("vldaft  loglik: %.3f\ngamlss2 loglik: %.3f\n",
            checks$gamlss2_loglik[["vldaft"]],
            checks$gamlss2_loglik[["gamlss2"]]))
```

We can also verify that Models 1, 3a, and 3b agree:

```{r gamlss2-check, eval = FALSE}
g1 <- gamlss2(
  Surv(YTCHD, CHD) ~ ln.Age. + ln.SBP. + ln.Chol. + MRW + Smoke,
  data = MLT50, family = AFT_Weibull())

g3a <- gamlss2(
  Surv(YTCHD, CHD) ~ ln.Age. + ln.SBP. + ln.Chol. + MRW + Smoke | 1 | 1 | 1,
  data = MLT50, family = AFT_Weibull(theta = 2),
  control = gamlss2_control(maxit = 200))

g3b <- gamlss2(
  Surv(YTCHD, CHD) ~ ln.Age. + ln.SBP. + ln.Chol. + MRW + Smoke |
                      ln.Age. + ln.SBP. + ln.Chol. + MRW + Smoke,
  data = MLT50, family = AFT_Weibull())
```

```{r gamlss2-table, eval = checks$has_gamlss2}
print(checks$gamlss2_check, row.names = FALSE)
```

## Interpretation

The significant negative $\theta_1 = -0.95$ means that as $\mu$
increases (i.e., as predicted log-survival time increases due to lower
risk), the dispersion $\sigma$ also increases. Equivalently, the
cumulative hazard ratio for two individuals shrinks toward unity
over longer follow-up. This implies:

- Neither the standard Weibull model nor the Cox proportional hazards
  model is adequate for these data.
- The association between risk factors and relative risk diminishes
  over time since measurement---a "converging hazards" phenomenon.

## Predicted Cumulative Incidence

Follow-up in the dataset ranges from less than 1 month to nearly 36
years (median 31 years, interquartile range 19--34 years). We restrict
predictions to 30 years to stay within the well-observed range.

To illustrate the nonproportional hazards effect, we compare predicted
cumulative incidence $F(t) = 1 - S(t)$ for two individuals differing
only in total cholesterol (160 vs 240 mg/dL), with all other covariates
held at their sample means.

Under the Weibull model, $S(t) = \exp\{-\exp(w)\}$ where
$w = (\log t - \mu)/\sigma$.

```{r predict-setup}
t_grid <- prediction$t_grid
inc1_240 <- prediction$incidence$model1_chol240
inc1_160 <- prediction$incidence$model1_chol160
inc2_240 <- prediction$incidence$model2_chol240
inc2_160 <- prediction$incidence$model2_chol160
diff1 <- prediction$excess$model1
diff2 <- prediction$excess$model2
```

```{r predict-plot, fig.width = 6, fig.height = 5}
plot(t_grid, inc1_240, type = "l",
     col = "firebrick", lwd = 2, lty = 1, ylim = c(0, 0.6),
     xlab = "Years since baseline", ylab = "Cumulative CHD incidence",
     main = "Predicted Incidence: Cholesterol 240 vs 160")
lines(t_grid, inc1_160,
      col = "steelblue", lwd = 2, lty = 1)
lines(t_grid, inc2_240,
      col = "firebrick", lwd = 2, lty = 2)
lines(t_grid, inc2_160,
      col = "steelblue", lwd = 2, lty = 2)
legend("topleft",
       legend = c("Chol = 240, Model 1 (LLAFT)",
                  "Chol = 160, Model 1 (LLAFT)",
                  "Chol = 240, Model 2 (VLDAFT)",
                  "Chol = 160, Model 2 (VLDAFT)"),
       col = c("firebrick", "steelblue", "firebrick", "steelblue"),
       lty = c(1, 1, 2, 2), lwd = 2, bty = "n", cex = 0.85)
```

The difference in cumulative incidence between cholesterol 240 and 160
quantifies how much additional risk the higher cholesterol confers over
time under each model.

```{r diff-plot, fig.width = 6, fig.height = 4.5}
plot(t_grid, diff1, type = "l", lwd = 2, lty = 1,
     col = "black", ylim = c(0, max(diff1, diff2) * 1.1),
     xlab = "Years since baseline",
     ylab = "Difference in cumulative incidence",
     main = "Excess Incidence: Cholesterol 240 vs 160")
lines(t_grid, diff2, lwd = 2, lty = 2, col = "black")
legend("topleft",
       legend = c("Model 1 (LLAFT)", "Model 2 (VLDAFT)"),
       lty = c(1, 2), lwd = 2, bty = "n")
```

Under both models, higher cholesterol is associated with higher cumulative incidence of CHD over time.
The difference in cumulative incidence also grows over time.
The observation that the difference is always greater with the VLDAFT model reflects the model's ability to capture nonproportional hazards.
