---
title: "Poisson-Mixture Cure Families for vldaft"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Poisson-Mixture Cure Families for vldaft}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(vldaft)
library(gglite)
has_gamlss2 <- requireNamespace("gamlss2", quietly = TRUE)
```

## Motivation

The baseline `vldaft` model is already useful for prediction because it
directly models subject-specific survival curves. What it cannot represent
is a **survival plateau**: long-run survival that levels off above zero.

For indications with a plausible cure fraction, the package now provides
`gamlss2` family constructors that wrap the existing accelerated failure
time baselines in a **Poisson-mixture cure model**. The idea is to keep the
baseline AFT machinery exactly as before, but replace the survival function
\(S_0(t)\) with

\[
S(t) = \exp\{-\tau (1 - S_0(t))\},
\qquad
\tau = -\log(p),
\qquad
p = \operatorname{logit}^{-1}(\nu).
\]

Here \(p\) is the cure probability and \(\nu\) is modeled with a standard
logit link.

## Available family constructors

The current implementation keeps the **architecture generic** over the
supported baseline survival distributions:

- `Cure_Weibull(theta = 0L)`
- `Cure_Logistic(theta = 0L)`
- `Cure_Normal(theta = 0L)`
- `Cure_Cauchy(theta = 0L)`
- `Cure_Gamma(theta = 0L, base_nu = 1)`
- `Cure_Exponential()`

For the cure families, the parameter blocks are:

- `mu`: location / linear predictor for the baseline AFT model
- `sigma`: log-scale parameter (except for `Cure_Exponential()`, where
  \(\sigma \equiv 1\))
- `nu`: cure probability on the response scale, with a logit link
- `theta1`, `theta2`, ...: optional VLDAFT coupling terms linking
  `log(sigma)` to `mu`

So a `gamlss2` call for a one-theta cure Weibull model uses the formula
layout `mu | sigma | nu | theta1`.

```{r gamlss2-cure-example, eval = FALSE}
library(gamlss2)
library(survival)

fit <- gamlss2(
  Surv(time, status) ~ x1 + x2 | 1 | x1 + x2 | 1,
  data = my_data,
  family = Cure_Weibull(theta = 1)
)
```

## Likelihood

Let \(F_0(t) = 1 - S_0(t)\) denote the baseline CDF, and let

\[
w_i = \frac{\log t_i - \mu_i}{\sigma_{i,\mathrm{eff}}}
\]

be the standardized AFT residual, with the same `theta`-polynomial
coupling used by the non-cure families:

\[
\log(\sigma_{i,\mathrm{eff}})
= \log(\sigma_i) + \sum_{k=1}^{K} \theta_k (\mu_i - \bar{\mu})^k.
\]

For an observed event, the cure-model density is

\[
f(t_i) = \tau_i f_0(t_i) \exp\{-\tau_i F_0(t_i)\},
\]

so the log-likelihood contribution is

\[
\ell_i^{(\mathrm{event})}
= \log(\tau_i) + \log f_0(t_i) - \tau_i F_0(t_i).
\]

For a right-censored observation,

\[
\ell_i^{(\mathrm{cens})} = -\tau_i F_0(t_i).
\]

For a left-censored observation known only to satisfy \(T_i \le a_i\),

\[
\ell_i^{(\mathrm{left})}
= \log\left(1 - \exp\{-\tau_i F_0(a_i)\}\right).
\]

For an interval-censored observation \(a_i < T_i \le b_i\),

\[
\ell_i^{(\mathrm{int})}
= \log\left[
    \exp\{-\tau_i F_0(a_i)\}
    - \exp\{-\tau_i F_0(b_i)\}
  \right].
\]

The ordinary non-cure AFT model uses the same censoring structure with
\(S_0(t)\) in place of \(\exp\{-\tau_i F_0(t)\}\). Thus, for an
interval-censored non-cure observation,

\[
\ell_i^{(\mathrm{int},0)}
= \log\left\{S_0(a_i) - S_0(b_i)\right\}.
\]

For left truncation at \(t_{0i}\), we condition on survival to entry and
add

\[
\tau_i F_0(t_{0i})
\]

to the contribution above.

## First partials

Let \(\alpha\) be any baseline parameter block (`mu`, `sigma`, `theta1`,
...). Because

\[
\frac{\partial F_0}{\partial \alpha}
  = -S_0 \frac{\partial \log S_0}{\partial \alpha},
\]

the cure-model score takes a simple form:

\[
\frac{\partial \ell_i}{\partial \alpha}
= I(\delta_i = 1)\frac{\partial \log f_0(t_i)}{\partial \alpha}
  + \tau_i S_0(t_i)\frac{\partial \log S_0(t_i)}{\partial \alpha},
\]

with the left-truncation correction

\[
-\tau_i S_0(t_{0i})
\frac{\partial \log S_0(t_{0i})}{\partial \alpha}.
\]

For the cure block itself, the implementation uses \(\nu\) as the additive
predictor with a logit link, so \(p = \operatorname{logit}^{-1}(\nu)\) and

\[
\frac{\partial \tau}{\partial \nu} = -(1 - p).
\]

Therefore

\[
\frac{\partial \ell_i^{(\mathrm{event})}}{\partial \nu}
= (1-p_i)\left(F_0(t_i) - \frac{1}{\tau_i}\right),
\qquad
\frac{\partial \ell_i^{(\mathrm{cens})}}{\partial \nu}
= (1-p_i)F_0(t_i),
\]

with a left-truncation subtraction of \((1-p_i)F_0(t_{0i})\).

For interval censoring it is useful to write

\[
C_i(t) = \exp\{-\tau_i F_0(t)\}.
\]

The baseline-parameter score for \(a_i < T_i \le b_i\) is then

\[
\frac{\partial \ell_i^{(\mathrm{int})}}{\partial \alpha}
=
\frac{
  C_i(a_i)\tau_i S_0(a_i)
    \frac{\partial \log S_0(a_i)}{\partial \alpha}
  -
  C_i(b_i)\tau_i S_0(b_i)
    \frac{\partial \log S_0(b_i)}{\partial \alpha}
}{
  C_i(a_i) - C_i(b_i)
}.
\]

For the cure block,

\[
\frac{\partial \ell_i^{(\mathrm{int})}}{\partial \nu}
=
\frac{
  C_i(a_i)(1-p_i)F_0(a_i)
  -
  C_i(b_i)(1-p_i)F_0(b_i)
}{
  C_i(a_i) - C_i(b_i)
}.
\]

## Second partials

The full Hessian is algebraically messy but follows directly from the score
expressions above. For the one-dimensional cure block, differentiating once
more gives

\[
\frac{\partial^2 \ell_i^{(\mathrm{event})}}{\partial \nu^2}
= -p_i(1-p_i)\left(F_0(t_i) - \frac{1}{\tau_i}\right)
   \frac{-(1-p_i)^2}{\tau_i^2},
\]

and for a censored observation,

\[
\frac{\partial^2 \ell_i^{(\mathrm{cens})}}{\partial \nu^2}
= -p_i(1-p_i)F_0(t_i),
\]

again with the obvious left-truncation correction. The standard non-cure
`vldaft()` C and Rust back ends use analytic observed-information updates,
including the interval-censored contribution
\(\log\{S_0(a_i) - S_0(b_i)\}\). The standalone cure optimizer uses the
compiled analytic score together with `optimHess()` for the covariance
matrix.

## A small simulation

The code below simulates from an intercept-only cure exponential model and
then shows the matching family constructor.

```{r cure-sim}
set.seed(21)
n <- 300
mu_true <- 1.3
p_true <- 0.25
tau <- -log(p_true)
u <- runif(n)
has_event <- u <= (1 - p_true)
t_event <- rep(Inf, n)
f0 <- -log1p(-u[has_event]) / tau
t_event[has_event] <- exp(mu_true + log(-log(1 - f0)))
c_time <- rexp(n, rate = 0.02)
time <- pmin(t_event, c_time)
status <- as.integer(t_event <= c_time)

mean(status == 0)
```

## Demo: `gamlss2` and standalone `vldaft()`

The cure families were built first at the `gamlss2` layer and then wired
into a standalone C-backed `vldaft(cure = TRUE)` path. On the same simulated
dataset, the two fits should achieve nearly the same objective value.

`vldaft()` now reports the linear blocks (`mu`, `sigma`, and `cure`) back on
the **raw covariate scale**, even when it internally centers covariates for
stability. The original training means are stored on the fit object and
reused inside `predict()`, so the printed coefficients stay easy to compare
with `gamlss2` while prediction remains faithful to the internally centered
fit.

```{r cure-demo-data}
set.seed(20260422)
n_demo <- 250
x1 <- rnorm(n_demo)
x2 <- rbinom(n_demo, 1, 0.4)
mu_demo <- 1.8 + 0.5 * x1 - 0.35 * x2
sigma_demo <- exp(-0.25)
nu_demo <- qlogis(0.22) - 0.35 * x1 + 0.55 * x2
p_demo <- plogis(nu_demo)
tau_demo <- -log(p_demo)
u_demo <- runif(n_demo)
has_event_demo <- u_demo <= (1 - p_demo)
t_event_demo <- rep(Inf, n_demo)
f0_demo <- -log1p(-u_demo[has_event_demo]) / tau_demo[has_event_demo]
t_event_demo[has_event_demo] <- exp(mu_demo[has_event_demo] +
  sigma_demo * log(-log(1 - f0_demo)))
c_time_demo <- rexp(n_demo, rate = 0.03)

demo_dat <- data.frame(
  time = pmin(t_event_demo, c_time_demo),
  status = as.integer(t_event_demo <= c_time_demo),
  x1 = x1,
  x2 = x2
)

mean(demo_dat$status == 0)
```

```{r cure-vldaft-fit}
fit_vldaft <- vldaft(
  survival::Surv(time, status) ~ x1 + x2 | 1 | x1 + x2,
  data = demo_dat,
  dist = "weibull",
  cure = TRUE,
  backend = "c"
)

coef(fit_vldaft)
fit_vldaft$loglik
```

```{r cure-gamlss2-fit, eval = has_gamlss2}
library(gamlss2)

fit_gamlss2 <- gamlss2(
  survival::Surv(time, status) ~ x1 + x2 | 1 | x1 + x2,
  data = demo_dat,
  family = Cure_Weibull(),
  control = gamlss2_control(maxit = 100)
)

coef(fit_gamlss2)
logLik(fit_gamlss2)
```

```{r cure-fit-compare, eval = has_gamlss2}
compare <- data.frame(
  term = c(
    "loglik",
    "scale_intercept",
    "mu_intercept",
    "mu_x1",
    "mu_x2",
    "cure_intercept",
    "cure_x1",
    "cure_x2"
  ),
  vldaft = c(
    fit_vldaft$loglik,
    unname(coef(fit_vldaft)["gamma:(Intercept)"]),
    unname(coef(fit_vldaft)["eta:(Intercept)"]),
    unname(coef(fit_vldaft)["eta:x1"]),
    unname(coef(fit_vldaft)["eta:x2"]),
    unname(coef(fit_vldaft)["cure:(Intercept)"]),
    unname(coef(fit_vldaft)["cure:x1"]),
    unname(coef(fit_vldaft)["cure:x2"])
  ),
  gamlss2 = c(
    as.numeric(logLik(fit_gamlss2)),
    unname(coef(fit_gamlss2)["sigma.p.(Intercept)"]),
    unname(coef(fit_gamlss2)["mu.p.(Intercept)"]),
    unname(coef(fit_gamlss2)["mu.p.x1"]),
    unname(coef(fit_gamlss2)["mu.p.x2"]),
    unname(coef(fit_gamlss2)["nu.p.(Intercept)"]),
    unname(coef(fit_gamlss2)["nu.p.x1"]),
    unname(coef(fit_gamlss2)["nu.p.x2"])
  )
)
compare$abs_diff <- abs(compare$vldaft - compare$gamlss2)
compare
```

```{r cure-fit-compare-predictions, eval = has_gamlss2}
coef_g <- coef(fit_gamlss2)

pred_compare <- data.frame(
  metric = c(
    "logLik difference",
    "cor(mu linear predictor)",
    "max abs diff mu",
    "cor(cure probability)",
    "max abs diff cure probability"
  ),
  value = c(
    abs(fit_vldaft$loglik - as.numeric(logLik(fit_gamlss2))),
    cor(
      predict(fit_vldaft, newdata = demo_dat, type = "lp"),
      coef_g[["mu.p.(Intercept)"]] +
        coef_g[["mu.p.x1"]] * demo_dat$x1 +
        coef_g[["mu.p.x2"]] * demo_dat$x2
    ),
    max(abs(
      predict(fit_vldaft, newdata = demo_dat, type = "lp") -
        (coef_g[["mu.p.(Intercept)"]] +
           coef_g[["mu.p.x1"]] * demo_dat$x1 +
           coef_g[["mu.p.x2"]] * demo_dat$x2)
    )),
    cor(
      predict(fit_vldaft, newdata = demo_dat, type = "cure"),
      plogis(
        coef_g[["nu.p.(Intercept)"]] +
          coef_g[["nu.p.x1"]] * demo_dat$x1 +
          coef_g[["nu.p.x2"]] * demo_dat$x2
      )
    ),
    max(abs(
      predict(fit_vldaft, newdata = demo_dat, type = "cure") -
        plogis(
          coef_g[["nu.p.(Intercept)"]] +
            coef_g[["nu.p.x1"]] * demo_dat$x1 +
            coef_g[["nu.p.x2"]] * demo_dat$x2
        )
    ))
  )
)
pred_compare
```

On this simulation, the two optimizers land at almost the same
log-likelihood, and the implied linear predictors and cure probabilities are
essentially identical. With `vldaft()` reporting coefficients back on the raw
covariate scale, the intercepts are also directly comparable to the
`gamlss2` coefficients.

```{r cure-gamlss2-unavailable, eval = !has_gamlss2}
cat("`gamlss2` is not installed, so only the standalone `vldaft()` cure fit",
    "is shown in this render.\n")
```

```{r cure-exp-family}
fam <- Cure_Exponential()
par <- list(
  mu = rep(mu_true, 3),
  nu = rep(p_true, 3)
)

fam$cure_probability(par)
fam$survival(c(1, 2, 5), par)
```

```{r cure-exp-plot}
t_grid <- seq(0.1, 18, length.out = 200)
plot_df <- data.frame(
  time = t_grid,
  survival = fam$survival(t_grid, list(mu = mu_true, nu = p_true))
)

g2(plot_df, survival ~ time, title = "Cure exponential survival curve") |>
  mark_line() |>
  scale_x(domain = c(0, 18), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time", values = c(0, 6, 12, 18)) |>
  axis_y(title = "Survival probability", values = c(0, 0.2, 0.4, 0.6, 0.8, 1))
```

## Notes

- The cure-layer architecture is distribution-generic, but initial
  validation focused on the Weibull / exponential cases.
- In this first pass, `nu` is a **linear regression block only**. A future
  extension could let `nu` depend on low-order powers of `mu`, mirroring the
  VLDAFT coupling already used for `sigma`.
- The Poisson-mixture cure model does **not** collapse to the plain baseline
  family by simply sending \(p \to 0\); that common intuition is incorrect
  for this construction, so tests in this package focus on score
  consistency, left-truncation handling, and parameter recovery instead.
