---
title: "Poisson-Mixture Cure Families for vldaft"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Poisson-Mixture Cure Families for vldaft}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(vldaft)
library(gglite)
```

## Motivation

The baseline `vldaft` model is already useful for prediction because it
directly models subject-specific survival curves. What it cannot represent
is a **survival plateau**: long-run survival that levels off above zero.

For indications with a plausible cure fraction, the package now provides
`gamlss2` family constructors that wrap the existing accelerated failure
time baselines in a **Poisson-mixture cure model**. The idea is to keep the
baseline AFT machinery exactly as before, but replace the survival function
\(S_0(t)\) with

\[
S(t) = \exp\{-\tau (1 - S_0(t))\},
\qquad
\tau = -\log(p),
\qquad
p = \operatorname{logit}^{-1}(\nu).
\]

Here \(p\) is the cure probability and \(\nu\) is modeled with a standard
logit link.

## Available family constructors

The current implementation keeps the **architecture generic** over the
supported baseline survival distributions:

- `Cure_Weibull(theta = 0L)`
- `Cure_Logistic(theta = 0L)`
- `Cure_Normal(theta = 0L)`
- `Cure_Cauchy(theta = 0L)`
- `Cure_Gamma(theta = 0L, base_nu = 1)`
- `Cure_Exponential()`

For the cure families, the parameter blocks are:

- `mu`: location / linear predictor for the baseline AFT model
- `sigma`: log-scale parameter (except for `Cure_Exponential()`, where
  \(\sigma \equiv 1\))
- `nu`: cure probability on the response scale, with a logit link
- `theta1`, `theta2`, ...: optional VLDAFT coupling terms linking
  `log(sigma)` to `mu`

So a `gamlss2` call for a one-theta cure Weibull model uses the formula
layout `mu | sigma | nu | theta1`.

```{r gamlss2-cure-example, eval = FALSE}
library(gamlss2)
library(survival)

fit <- gamlss2(
  Surv(time, status) ~ x1 + x2 | 1 | x1 + x2 | 1,
  data = my_data,
  family = Cure_Weibull(theta = 1)
)
```

## Likelihood

Let \(F_0(t) = 1 - S_0(t)\) denote the baseline CDF, and let

\[
w_i = \frac{\log t_i - \mu_i}{\sigma_{i,\mathrm{eff}}}
\]

be the standardized AFT residual, with the same `theta`-polynomial
coupling used by the non-cure families:

\[
\log(\sigma_{i,\mathrm{eff}})
= \log(\sigma_i) + \sum_{k=1}^{K} \theta_k (\mu_i - \bar{\mu})^k.
\]

For an observed event, the cure-model density is

\[
f(t_i) = \tau_i f_0(t_i) \exp\{-\tau_i F_0(t_i)\},
\]

so the log-likelihood contribution is

\[
\ell_i^{(\mathrm{event})}
= \log(\tau_i) + \log f_0(t_i) - \tau_i F_0(t_i).
\]

For a right-censored observation,

\[
\ell_i^{(\mathrm{cens})} = -\tau_i F_0(t_i).
\]

For left truncation at \(t_{0i}\), we condition on survival to entry and
add

\[
\tau_i F_0(t_{0i})
\]

to the contribution above.

## First partials

Let \(\alpha\) be any baseline parameter block (`mu`, `sigma`, `theta1`,
...). Because

\[
\frac{\partial F_0}{\partial \alpha}
  = -S_0 \frac{\partial \log S_0}{\partial \alpha},
\]

the cure-model score takes a simple form:

\[
\frac{\partial \ell_i}{\partial \alpha}
= I(\delta_i = 1)\frac{\partial \log f_0(t_i)}{\partial \alpha}
  \tau_i S_0(t_i)\frac{\partial \log S_0(t_i)}{\partial \alpha},
\]

with the left-truncation correction

\[
-\tau_i S_0(t_{0i})
\frac{\partial \log S_0(t_{0i})}{\partial \alpha}.
\]

For the cure block itself, the implementation uses \(\nu\) as the additive
predictor with a logit link, so \(p = \operatorname{logit}^{-1}(\nu)\) and

\[
\frac{\partial \tau}{\partial \nu} = -(1 - p).
\]

Therefore

\[
\frac{\partial \ell_i^{(\mathrm{event})}}{\partial \nu}
= (1-p_i)\left(F_0(t_i) - \frac{1}{\tau_i}\right),
\qquad
\frac{\partial \ell_i^{(\mathrm{cens})}}{\partial \nu}
= (1-p_i)F_0(t_i),
\]

with a left-truncation subtraction of \((1-p_i)F_0(t_{0i})\).

## Second partials

The full Hessian is algebraically messy but follows directly from the score
expressions above. For the one-dimensional cure block, differentiating once
more gives

\[
\frac{\partial^2 \ell_i^{(\mathrm{event})}}{\partial \nu^2}
= -p_i(1-p_i)\left(F_0(t_i) - \frac{1}{\tau_i}\right)
   \frac{-(1-p_i)^2}{\tau_i^2},
\]

and for a censored observation,

\[
\frac{\partial^2 \ell_i^{(\mathrm{cens})}}{\partial \nu^2}
= -p_i(1-p_i)F_0(t_i),
\]

again with the obvious left-truncation correction. The current
implementation exposes **analytic scores**; a fully analytic Hessian for
every parameter block is a natural next extension if optimization speed
becomes limiting.

## A small simulation

The code below simulates from an intercept-only cure exponential model and
then shows the matching family constructor.

```{r cure-sim}
set.seed(21)
n <- 300
mu_true <- 1.3
p_true <- 0.25
tau <- -log(p_true)
u <- runif(n)
has_event <- u <= (1 - p_true)
t_event <- rep(Inf, n)
f0 <- -log1p(-u[has_event]) / tau
t_event[has_event] <- exp(mu_true + log(-log(1 - f0)))
c_time <- rexp(n, rate = 0.02)
time <- pmin(t_event, c_time)
status <- as.integer(t_event <= c_time)

mean(status == 0)
```

```{r cure-exp-family}
fam <- Cure_Exponential()
par <- list(
  mu = rep(mu_true, 3),
  nu = rep(p_true, 3)
)

fam$cure_probability(par)
fam$survival(c(1, 2, 5), par)
```

```{r cure-exp-plot}
t_grid <- seq(0.1, 20, length.out = 200)
plot_df <- data.frame(
  time = t_grid,
  survival = fam$survival(t_grid, list(mu = mu_true, nu = p_true))
)

g2(plot_df, survival ~ time, title = "Cure exponential survival curve") |>
  mark_line() |>
  axis_x(title = "Time") |>
  axis_y(title = "Survival probability")
```

## Notes

- The cure-layer architecture is distribution-generic, but initial
  validation focused on the Weibull / exponential cases.
- In this first pass, `nu` is a **linear regression block only**. A future
  extension could let `nu` depend on low-order powers of `mu`, mirroring the
  VLDAFT coupling already used for `sigma`.
- The Poisson-mixture cure model does **not** collapse to the plain baseline
  family by simply sending \(p \to 0\); that common intuition is incorrect
  for this construction, so tests in this package focus on score
  consistency, left-truncation handling, and parameter recovery instead.
