---
title: "Motivating the Weibull VLDAFT and Cure Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Motivating the Weibull VLDAFT and Cure Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

library(gglite)
```

## Introduction

This vignette provides a visual, equation-driven introduction to the
Weibull accelerated failure time (AFT) model, its VLDAFT extension
(Anderson, 1991), and the Poisson-mixture cure model implemented in
`vldaft`. The goal is to build intuition about how each parameter
shapes the survival and hazard functions before fitting real data.

## The Weibull AFT model

For a positive event time $T$ and covariates $\mathbf{x}$:

$$
\log T = \mu + \sigma\,\varepsilon, \qquad \varepsilon \sim \text{Gumbel (standard extreme value)}
$$

The survival function is:

$$
S(t) = \exp\!\left\{-\exp\!\left(\frac{\log t - \mu}{\sigma}\right)\right\}
$$

Two parameters control the shape of this curve:
 
- **Location** $\mu$: the expected log-time. Larger $\mu$ shifts the
  survival curve to the right — subjects live longer.
- **Scale** $\sigma$: controls the spread of event times. The Weibull
  shape parameter is $\alpha = 1/\sigma$.

When $\sigma$ is constant across subjects, the Weibull AFT model is
equivalent to a **proportional hazards** model with constant hazard ratio.

## Helper functions

```{r helpers}
weibull_S <- function(t, mu, sigma) {
  exp(-exp((log(t) - mu) / sigma))
}

weibull_h <- function(t, mu, sigma) {
  alpha <- 1 / sigma
  lambda <- exp(-mu / sigma)
  alpha * lambda * t^(alpha - 1)
}

cure_S <- function(t, mu, sigma, p) {
  S0 <- weibull_S(t, mu, sigma)
  tau <- -log(p)
  exp(-tau * (1 - S0))
}

t_grid <- seq(0.01, 30, length.out = 300)
```

## Effect of location ($\mu$)

Holding $\sigma = 0.8$ fixed, we vary $\mu$ to see the shift effect:

```{r location-effect}
mu_vals <- c(0.5, 1.0, 1.5, 2.0, 2.5, 3.0)
sigma_fixed <- 0.8

df <- do.call(rbind, lapply(mu_vals, function(mu) {
  data.frame(time = t_grid,
             survival = weibull_S(t_grid, mu, sigma_fixed),
             mu = paste0("mu = ", mu))
}))

g2(df, survival ~ time, color = "mu",
   title = "Weibull survival: varying location (sigma = 0.8)") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "S(t)")
```

Each unit increase in $\mu$ corresponds to a multiplicative change in the
median survival time of $\exp(1) \approx 2.7$ (holding $\sigma$ fixed).
Covariates enter the model through $\mu = \boldsymbol\beta'\mathbf{x}$,
so their effect is to **accelerate or decelerate** the time to event.

## Effect on the hazard function

```{r location-hazard}
df_h <- do.call(rbind, lapply(c(1.0, 2.0, 3.0), function(mu) {
  data.frame(time = t_grid,
             hazard = weibull_h(t_grid, mu, sigma_fixed),
             mu = paste0("mu = ", mu))
}))

g2(df_h, hazard ~ time, color = "mu",
   title = "Weibull hazard: varying location (sigma = 0.8)") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "h(t)")
```

Under constant $\sigma$, the hazard functions are related by a constant
ratio — this is the proportional hazards property. The ratio
$h_A(t)/h_B(t)$ does not depend on $t$.

## Effect of scale ($\sigma$)

Now hold $\mu = 2.0$ fixed and vary $\sigma$:

```{r scale-effect}
sigma_vals <- c(0.3, 0.5, 0.8, 1.0, 1.5)

df_s <- do.call(rbind, lapply(sigma_vals, function(s) {
  data.frame(time = t_grid,
             survival = weibull_S(t_grid, 2.0, s),
             sigma = paste0("sigma = ", s))
}))

g2(df_s, survival ~ time, color = "sigma",
   title = "Weibull survival: varying scale (mu = 2.0)") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "S(t)")
```

The scale parameter controls the **shape** of the hazard:

- $\sigma < 1$ ($\alpha > 1$): increasing hazard — aging, wear-out,
  accumulating damage.
- $\sigma = 1$ ($\alpha = 1$): constant hazard — the exponential distribution.
- $\sigma > 1$ ($\alpha < 1$): decreasing hazard — early risk followed
  by a selection effect or healing.

```{r scale-hazard}
df_sh <- do.call(rbind, lapply(c(0.5, 0.8, 1.0, 1.5), function(s) {
  h <- weibull_h(t_grid, 2.0, s)
  h[h > 2] <- NA
  data.frame(time = t_grid, hazard = h,
             sigma = paste0("sigma = ", s))
}))
df_sh <- df_sh[!is.na(df_sh$hazard), ]

g2(df_sh, hazard ~ time, color = "sigma",
   title = "Weibull hazard: varying scale (mu = 2.0)") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "h(t)")
```

## Combining location and scale

When both parameters vary across subjects, we get a rich family of
survival curves:

```{r combined}
combos <- list(
  list(mu = 1.0, sigma = 0.5, label = "mu=1.0, sigma=0.5 (high risk, steep)"),
  list(mu = 2.0, sigma = 0.5, label = "mu=2.0, sigma=0.5 (low risk, steep)"),
  list(mu = 1.0, sigma = 1.2, label = "mu=1.0, sigma=1.2 (high risk, gradual)"),
  list(mu = 2.0, sigma = 1.2, label = "mu=2.0, sigma=1.2 (low risk, gradual)")
)

df_c <- do.call(rbind, lapply(combos, function(p) {
  data.frame(time = t_grid,
             survival = weibull_S(t_grid, p$mu, p$sigma),
             profile = p$label)
}))

g2(df_c, survival ~ time, color = "profile",
   title = "Weibull survival: four location-scale combinations") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "S(t)")
```

This motivates the idea that different subjects may have different
*shapes* of their hazard function, not just different levels.

## The limitation of constant $\sigma$

In the standard Weibull AFT model, $\sigma$ is a single global parameter
shared by all subjects. This implies:

$$
\frac{h_A(t)}{h_B(t)} = \text{constant for all } t
$$

In practice, hazard ratios often:

- **Converge** toward 1 over time (risk-factor effects wane as both groups
  age or develop competing risks)
- **Diverge** (delayed treatment effects, threshold phenomena)
- **Cross** (early harm, late benefit — e.g., aggressive surgery vs.
  conservative management)

A model with constant $\sigma$ cannot represent any of these patterns.

## The VLDAFT extension

Anderson (1991) proposed a simple, parsimonious generalization: let
$\sigma$ depend on the fitted location through a low-order polynomial:

$$
\log \sigma_i = \gamma_0 + \theta_1 (\mu_i - \bar\mu)
$$

This adds **one** parameter ($\theta_1$) to the standard model and
breaks proportional hazards:

- $\theta_1 = 0$: standard AFT — proportional hazards.
- $\theta_1 < 0$: subjects with longer expected survival have *smaller*
  dispersion → hazards **converge** over time.
- $\theta_1 > 0$: hazards **diverge** over time.

### Proportional hazards ($\theta_1 = 0$)

```{r theta-zero}
mu_low <- 1.5; mu_high <- 2.5; mubar <- 2.0; gamma0 <- log(0.6)

make_theta_data <- function(th) {
  sig_low  <- exp(gamma0 + th * (mu_low  - mubar))
  sig_high <- exp(gamma0 + th * (mu_high - mubar))
  rbind(
    data.frame(time = t_grid,
               survival = weibull_S(t_grid, mu_low, sig_low),
               group = "high risk (mu = 1.5)"),
    data.frame(time = t_grid,
               survival = weibull_S(t_grid, mu_high, sig_high),
               group = "low risk (mu = 2.5)")
  )
}

g2(make_theta_data(0), survival ~ time, color = "group",
   title = "theta1 = 0: proportional hazards") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "S(t)")
```

The gap between curves remains constant on the log-cumulative-hazard
scale — this is proportional hazards.

### Converging hazards ($\theta_1 = -0.3$)

```{r theta-neg}
g2(make_theta_data(-0.3), survival ~ time, color = "group",
   title = "theta1 = -0.3: converging hazards") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "S(t)")
```

At late times, the curves are **closer together** — the initial
difference in risk wanes. This is the phenomenon Anderson (1991) found
in the Framingham Heart Study.

### Diverging hazards ($\theta_1 = 0.3$)

```{r theta-pos}
g2(make_theta_data(0.3), survival ~ time, color = "group",
   title = "theta1 = +0.3: diverging hazards") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "S(t)")
```

The gap between curves **widens** — a delayed or amplifying treatment
effect.

## Anderson (1991): the Framingham Heart Study

Anderson applied the VLDAFT model to 1,433 men aged 30–49 from the
Framingham Heart Study, followed for approximately 32 years for coronary
heart disease (CHD) incidence. The covariates were log(age), log(SBP),
log(cholesterol), Metropolitan relative weight, and a smoking indicator.

### Key results

| Model | Description | Parameters | Log-likelihood |
|:-----:|:-----------:|:----------:|:--------------:|
| 1 | Standard Weibull AFT | 7 | −4414.5 |
| 2 | VLDAFT ($\theta_1$) | 8 | −4404.0 |
| 3a | VLDAFT ($\theta_1, \theta_2$) | 9 | −4403.9 |
| 3b | Full scale regression | 12 | −4404.0 |

The likelihood ratio test for Model 2 vs Model 1 gives
$\chi^2_1 = 20.9$ ($p < 0.001$), strongly rejecting proportional hazards.
The estimated $\hat\theta_1 = -0.95$ indicates that
dispersion *increases* with predicted survival time — risk-factor effects
on the hazard converge over the 32-year follow-up.

Importantly, Models 3a and 3b (with 1–5 additional parameters) provide
**no improvement** over Model 2. A single parameter $\theta_1$ captures
the entire nonproportional hazards structure.

### Practical interpretation

In a standard proportional hazards model, a subject with high cholesterol
has an elevated hazard that persists unchanged for 32 years. The VLDAFT
model reveals that this elevated hazard **wanes** over time — perhaps
because surviving subjects develop competing risks (cancer, accidents),
or because the original risk factor measurement becomes less predictive
as it ages.

## The Poisson-mixture cure model

Some diseases (certain cancers, transplant outcomes, chronic infections)
show a **survival plateau**: a fraction of subjects who will never
experience the event. No amount of location or scale adjustment in a
standard AFT model can produce a flat tail.

The Poisson-mixture cure model wraps the baseline survival:

$$
S(t) = \exp\{-\tau(1 - S_0(t))\}, \qquad \tau = -\log(p)
$$

where $p$ is the **cure probability** and $S_0(t)$ is the Weibull
baseline. As $t \to \infty$, $S_0(t) \to 0$, so
$S(\infty) = \exp(-\tau) = p$ — a plateau at the cure rate.

### Effect of the cure rate

Fix $\mu = 2.0$ and $\sigma = 0.8$, vary $p$:

```{r cure-rate}
cure_rates <- c(0, 0.1, 0.3, 0.5)
df_cure <- do.call(rbind, lapply(cure_rates, function(p) {
  S <- if (p == 0) weibull_S(t_grid, 2.0, 0.8)
       else cure_S(t_grid, 2.0, 0.8, p)
  data.frame(time = t_grid, survival = S,
             cure = paste0("p = ", sprintf("%.0f%%", p * 100)))
}))

g2(df_cure, survival ~ time, color = "cure",
   title = "Effect of cure rate on Weibull survival (mu = 2, sigma = 0.8)") |>
  mark_line() |>
  scale_x(domain = c(0, 30), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time") |>
  axis_y(title = "S(t)")
```

The plateau at $S(\infty) = p$ is clearly visible. The higher the cure
rate, the earlier the survival curve levels off.

### Effect of the shape parameter under a fixed cure rate

Now fix $\mu = \log(2) \approx 0.69$ (so that uncured subjects have
median event time around 2 years) and cure rate $p = 0.30$. Varying
$\sigma$ changes how quickly the curve descends toward the plateau:

```{r cure-shape}
mu_fixed <- log(2)
p_fixed <- 0.3
sigma_cure_vals <- c(0.3, 0.5, 0.8, 1.0, 1.5)
t_grid5 <- seq(0.01, 5, length.out = 300)

df_cs <- do.call(rbind, lapply(sigma_cure_vals, function(s) {
  data.frame(
    time = t_grid5,
    survival = cure_S(t_grid5, mu_fixed, s, p_fixed),
    sigma = paste0("sigma = ", s, " (alpha = ", round(1/s, 2), ")")
  )
}))

g2(df_cs, survival ~ time, color = "sigma",
   title = "Cure Weibull: varying shape (mu = log(2), cure = 30%)") |>
  mark_line() |>
  scale_x(domain = c(0, 5), nice = FALSE) |>
  scale_y(domain = c(0, 1), nice = FALSE) |>
  axis_x(title = "Time (years)") |>
  axis_y(title = "S(t)")
```

All curves converge to $S(\infty) = 0.30$, but:

- **Small $\sigma$** (large $\alpha$): rapid initial decline, quick
  approach to the plateau. Most events happen in a concentrated window.
- **Large $\sigma$** (small $\alpha$): gradual, drawn-out decline.
  Events are spread over a longer period before the plateau is reached.

This distinction matters clinically: a steep curve with a plateau looks
very different from a gradual decline in a Kaplan–Meier plot, and the
treatment-planning implications differ substantially.

## Summary

| Parameter | Role | Key effect |
|-----------|------|------------|
| $\mu$ (location) | Linear predictor on log-time | Shifts survival curve left/right |
| $\sigma$ (scale) | Controls hazard shape | $\sigma < 1$: increasing hazard; $\sigma > 1$: decreasing |
| $\theta_1$ (coupling) | Links $\log\sigma$ to $\mu$ | Breaks proportional hazards with one parameter |
| $p$ (cure rate) | Long-run survival plateau | $S(\infty) = p$; adds a cured fraction |

The `vldaft` package implements all of these in a unified framework:

- `vldaft()` fits standard and VLDAFT models with C or Rust backends.
- `AFT_Weibull()`, `Cure_Weibull()`, etc. provide `gamlss2` families
  for flexible regression on each parameter.
- Right, left, and interval censoring plus left truncation are supported.

## References

- Anderson, K. M. (1991). A nonproportional hazards Weibull accelerated
  failure time regression model. *Biometrics*, **47**(1), 281–288.
- Rodrigues, J., Cancho, V. G., de Castro, M. & Louzada-Neto, F.
  (2009). On the unification of long-term survival models.
  *Statistics & Probability Letters*, **79**(6), 753–759.
